---
title: "Potyvirus Phylogenetics and CI Region Inference"
subtitle: "MAFFT alignment, codon-aware tree inference, and convergent evolution testing"
author: "Christina Muedsam"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
execute:
  echo: true
  warning: false
  error: false
---

<!-- SOURCE: 2026_CopilotPoty/potyvirus_pipeline_nogenbankr_final.qmd (2026-05-02) -->
<!-- NOTE: eval: false — requires rentrez and external files; remove once renv is synced -->
<!-- NOTE: external tools required: mafft, iqtree2 (run via system() calls) -->

# Introduction

This chapter performs a comprehensive phylogenetic pipeline for potyviruses,
without the Bioconductor `genbankr` package. It reads a MAFFT protein alignment,
separates references and isolates, retrieves reference CDS via `rentrez`, parses
CDS from plain GenBank flatfiles, projects to codon alignment, infers the CI
region automatically, extracts a short motif of interest, visualises these motifs
on the resulting tree, and tests for convergent evolution versus shared ancestry.

# Setup and Helpers

```{r setup}
#| include: false
library(Biostrings)
library(rentrez)
library(stringr)
library(dplyr)
library(tibble)
library(ape)
library(phytools)
library(ggplot2)
library(ggtree)
library(ggseqlogo)
library(readr)
```

Helper functions below implement GenBank parsing, accession extraction, codon
projection, safe translation etc.

```{r helpers}
has_genbank_id <- function(x) {
  grepl("[A-Z]{1,2}_?[0-9]{5,8}", x)
}

extract_accessions <- function(x) {
  unlist(stringr::str_extract_all(x,
    "[A-Z]{1,2}_?[0-9]{5,8}(?:\\.[0-9]+)?"))
}

normalize_name <- function(x) {
  tolower(gsub("[^a-z0-9]", "", x))
}

aa_similarity <- function(x, y) {
  n <- min(nchar(x), nchar(y))
  sum(strsplit(substr(x, 1, n), "")[[1]] == strsplit(substr(y, 1, n), "")[[1]]) / n
}

parse_cds_features <- function(gb_txt) {
  feat_start <- grep("^FEATURES", gb_txt)
  seq_start  <- grep("^ORIGIN", gb_txt)
  if (length(feat_start) == 0 || length(seq_start) == 0) return(list())
  feat_lines <- gb_txt[(feat_start + 1):(seq_start - 1)]
  cds_idx <- grep("^ +CDS +", feat_lines)
  if (length(cds_idx) == 0) return(list())
  feats <- list()
  for (i in cds_idx) {
    lines <- feat_lines[i]
    j <- i + 1
    while (j <= length(feat_lines) && grepl("^ {21}", feat_lines[j])) {
      lines <- c(lines, feat_lines[j]); j <- j + 1
    }
    loc <- sub("^ +CDS +", "", lines[1])
    product <- NA; prot_id <- NA
    for (l in lines) {
      if (grepl("/product=", l)) {
        m <- regmatches(l, regexec('"([^"]+)"', l))[[1]]
        if (length(m) >= 2) product <- m[2]
      }
      if (grepl("/protein_id=", l)) {
        m <- regmatches(l, regexec('"([^"]+)"', l))[[1]]
        if (length(m) >= 2) prot_id <- m[2]
      }
    }
    feats[[length(feats) + 1]] <- list(location = loc, product = product, protein_id = prot_id)
  }
  feats
}

extract_genbank_sequence <- function(gb_txt) {
  origin_start <- grep("^ORIGIN", gb_txt)
  if (length(origin_start) == 0) return(DNAString(""))
  seq_lines <- gb_txt[(origin_start + 1):length(gb_txt)]
  seq <- paste(gsub("[^acgtACGT]", "", seq_lines), collapse = "")
  DNAString(seq)
}

parse_location <- function(loc) {
  loc_clean <- gsub("complement\\(|join\\(|\\)", "", loc)
  parts <- unlist(strsplit(loc_clean, ","))
  coords <- lapply(parts, function(p) {
    nums <- as.integer(unlist(strsplit(p, "\\.\\.")))
    c(start = nums[1], end = nums[2])
  })
  do.call(rbind, coords)
}

extract_cds_sequence <- function(genome, cds_feat) {
  coords <- parse_location(cds_feat$location)
  seqs <- apply(coords, 1, function(x) subseq(genome, start = x["start"], end = x["end"]))
  do.call(xscat, as.list(seqs))
}

codon_align_from_protein <- function(prot_aln, cds) {
  out <- DNAStringSet()
  for (id in intersect(names(prot_aln), names(cds))) {
    prot   <- strsplit(as.character(prot_aln[[id]]), "")[[1]]
    codons <- substring(as.character(cds[[id]]),
                        seq(1, length(cds[[id]]), 3),
                        seq(3, length(cds[[id]]), 3))
    idx <- 1
    seq_out <- character()
    for (aa in prot) {
      if (aa == "-") {
        seq_out <- c(seq_out, "---")
      } else {
        if (idx > length(codons)) {
          seq_out <- c(seq_out, "NNN")
        } else {
          seq_out <- c(seq_out, codons[idx]); idx <- idx + 1
        }
      }
    }
    out[[id]] <- DNAString(paste(seq_out, collapse = ""))
  }
  out
}

write_iqtree_codon_partitions <- function(codon_aln, file = "iqtree_codon_partitions.txt") {
  aln_len <- width(codon_aln)[1]
  lines <- c(
    sprintf("DNA, codon1 = 1-%d\\3", aln_len),
    sprintf("DNA, codon2 = 2-%d\\3", aln_len),
    sprintf("DNA, codon3 = 3-%d\\3", aln_len)
  )
  writeLines(lines, file)
}

safe_translate <- function(seq) {
  s <- gsub("-", "", as.character(seq))
  trim_len <- nchar(s) - (nchar(s) %% 3)
  s <- substr(s, 1, trim_len)
  translate(DNAString(s))
}
```

# Data Input

```{r data-input}
prot_file <- here::here("data", "OPSR.Poty.Fig3.v16_align_withblasthits_protmafftadd.txt")
orf_file  <- here::here("data", "isolates_longest_orfs.fna")
cache_dir <- here::here("cache", "genbank")

prot_aln <- readAAStringSet(prot_file)
stopifnot(length(unique(width(prot_aln))) == 1)
cat("Protein alignment loaded: length =", width(prot_aln)[1], "aa\n")
```

# Reference and Isolate Separation

```{r separation}
ref_idx  <- has_genbank_id(names(prot_aln))
prot_ref <- prot_aln[ref_idx]
prot_orf <- prot_aln[!ref_idx]
cat(length(prot_ref), "reference sequences and", length(prot_orf), "isolate sequences.\n")
```

# Retrieve Reference CDS

This section downloads GenBank flatfiles for references and extracts the
polyprotein CDS via our parser.

```{r retrieve-cds}
#| eval: false
# Requires internet access + rentrez; run manually and cache results to data/
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
cds_ref  <- DNAStringSet()
ref_diag <- tibble(name = names(prot_ref), status = NA_character_, note = NA_character_)

for (i in seq_along(prot_ref)) {
  nm <- names(prot_ref)[i]; accs <- extract_accessions(nm)
  found <- FALSE
  for (acc in accs) {
    gb_file <- file.path(cache_dir, paste0(acc, ".gb"))
    if (!file.exists(gb_file)) {
      try({
        gb_txt <- entrez_fetch("nuccore", acc, rettype = "gb", retmode = "text")
        writeLines(gb_txt, gb_file)
      }, silent = TRUE)
    }
    if (!file.exists(gb_file)) next
    gb_txt <- readLines(gb_file)
    feats  <- parse_cds_features(gb_txt)
    if (length(feats) == 0) next
    genome <- extract_genbank_sequence(gb_txt)
    hit <- NULL
    for (cf in feats) {
      if (!is.na(cf$protein_id) && grepl(cf$protein_id, nm, fixed = TRUE)) { hit <- cf; break }
    }
    if (is.null(hit)) {
      target <- normalize_name(nm)
      for (cf in feats) {
        prod <- normalize_name(cf$product)
        if (!is.na(prod) && (prod == target || grepl(target, prod, fixed = TRUE))) { hit <- cf; break }
      }
    }
    if (is.null(hit)) {
      best <- -Inf; best_cf <- NULL
      for (cf in feats) {
        cds_seq_tmp <- extract_cds_sequence(genome, cf)
        aa  <- safe_translate(cds_seq_tmp)
        sim <- aa_similarity(as.character(aa), gsub("-", "", as.character(prot_ref[[nm]])))
        if (sim > best) { best <- sim; best_cf <- cf }
      }
      if (!is.null(best_cf) && best >= 0.7) { hit <- best_cf }
    }
    if (is.null(hit)) next
    cds_seq <- extract_cds_sequence(genome, hit)
    if (length(cds_seq) %% 3 != 0)
      cds_seq <- subseq(cds_seq, start = 1, end = length(cds_seq) - (length(cds_seq) %% 3))
    if (length(cds_seq) == 0) next
    cds_ref[[nm]] <- cds_seq
    ref_diag$status[i] <- "ok"
    ref_diag$note[i]   <- acc
    found <- TRUE
    break
  }
  if (!found) {
    ref_diag$status[i] <- "failed"; ref_diag$note[i] <- "no CDS"
  }
}
```

# Read Isolate CDS

```{r read-isolate-cds}
#| eval: false
cds_orf <- readDNAStringSet(orf_file)
cds_orf <- DNAStringSet(lapply(cds_orf, function(x) {
  len <- length(x)
  if (len %% 3 != 0) x <- subseq(x, start = 1, end = len - (len %% 3))
  x
}))
```

# Merge and Project to Codon Alignment

```{r codon-projection}
#| eval: false
cds_all  <- c(cds_ref, cds_orf)
keep_ids <- intersect(names(prot_aln), names(cds_all))
codon_aln <- codon_align_from_protein(prot_aln[keep_ids], cds_all)
stopifnot(length(unique(width(codon_aln))) == 1)
writeXStringSet(codon_aln, here::here("data", "potyvirus_codon_alignment.fna"))
cat("Codon alignment constructed: length =", width(codon_aln)[1], "nt\n")
```

# Partition Files for IQ-TREE

```{r partitions}
#| eval: false
write_iqtree_codon_partitions(codon_aln,
  file = here::here("data", "iqtree_codon_partitions.txt"))

potyvirus_subproteins <- tibble(
  protein  = c("NIb", "CP"),
  aa_start = c(4100, 5600),
  aa_end   = c(5600, 6619)
) %>% mutate(nt_start = (aa_start - 1) * 3 + 1, nt_end = aa_end * 3)

lines_sub <- c()
for (i in seq_len(nrow(potyvirus_subproteins))) {
  p <- potyvirus_subproteins$protein[i]
  s <- potyvirus_subproteins$nt_start[i]
  e <- potyvirus_subproteins$nt_end[i]
  lines_sub <- c(lines_sub,
    sprintf("DNA, %s_codon1 = %d-%d\\3", p, s, e),
    sprintf("DNA, %s_codon2 = %d-%d\\3", p, s + 1, e),
    sprintf("DNA, %s_codon3 = %d-%d\\3", p, s + 2, e))
}
writeLines(lines_sub, here::here("data", "iqtree_subprotein_codon_partitions.txt"))
```

# Automatic CI Coordinates Inference

We infer the CI region from GenBank annotations where available and provide a
fallback region.

```{r infer-ci}
#| eval: false
ci_start <- NA; ci_end <- NA
for (nm in names(prot_ref)) {
  accs <- extract_accessions(nm)
  if (length(accs) == 0) next
  gb_file <- file.path(cache_dir, paste0(accs[1], ".gb"))
  if (!file.exists(gb_file)) next
  gb_txt  <- readLines(gb_file)
  feats   <- parse_cds_features(gb_txt)
  ci_feats <- Filter(function(x) grepl("cylindrical inclusion|\\bci\\b", tolower(x$product)), feats)
  if (length(ci_feats) > 0) {
    coords   <- parse_location(ci_feats[[1]]$location)
    ci_start <- min(coords[, "start"]) / 3
    ci_end   <- max(coords[, "end"]) / 3
    break
  }
}
if (is.na(ci_start) || is.na(ci_end)) {
  aa_len   <- width(prot_aln)[1]
  ci_start <- floor(aa_len * 2 / 7)
  ci_end   <- ceiling(aa_len * 3 / 7)
}
ci_nt_start <- (ceiling(ci_start) - 1) * 3 + 1
ci_nt_end   <- ceiling(ci_end) * 3
cat(sprintf("CI region: aa %d-%d; nt %d-%d\n", ci_start, ci_end, ci_nt_start, ci_nt_end))
```

# Extract CI Region

```{r extract-ci}
#| eval: false
codon_CI <- subseq(codon_aln, start = ci_nt_start, end = ci_nt_end)
writeXStringSet(codon_CI, here::here("data", "potyvirus_CI_codon_alignment.fna"))
cat("CI alignment length =", width(codon_CI)[1], "nt\n")
```

# Motif Definition and Extraction

Define your motif below and extract its presence/position.

```{r motif-extraction}
#| eval: false
motif_nt <- "ATGGACTTCGACGTTGAA"   # 18-nt motif in CI region
motif_aa <- as.character(translate(DNAString(motif_nt)))

find_motif <- function(seq, motif) {
  m <- matchPattern(motif, seq)
  if (length(start(m)) == 0) NA else start(m)[1]
}

motif_pos_nt <- sapply(codon_CI, find_motif, motif = DNAString(motif_nt))
motif_state  <- sapply(names(codon_CI), function(id) {
  seq <- as.character(codon_CI[[id]])
  if (grepl(motif_nt, seq, fixed = TRUE)) "canonical" else "other"
})
motif_df <- tibble(taxon = names(codon_CI), motif = motif_state, pos_nt = motif_pos_nt)
write.csv(motif_df, here::here("data", "CI_motif_states.csv"), row.names = FALSE)
```

# Phylogeny on CI Region

Infer the phylogeny using IQ-TREE, or read a precomputed tree.

```{r tree-ci}
# Run this command in a terminal to infer the CI tree:
# iqtree2 -s potyvirus_CI_codon_alignment.fna -p iqtree_codon_partitions.txt -m MFP -B 1000 -T AUTO -pre potyvirus_CI
# ci_tree <- read.tree(here::here("data", "potyvirus_CI.treefile"))
```

# Visualisation: Trait Mapping

```{r plot-tree}
#| fig-width: 9
#| fig-height: 9
#| fig-cap: "Cylindrical-inclusion (CI) protein phylogeny (circular layout), tips coloured by genus. Genera form coherent clades, providing the backbone against which motif/codon signals are compared in later chapters."
# Load a precomputed CI tree. The IQ-TREE codon tree needs rentrez + IQ-TREE;
# scripts/build_ci_tree.R produces an equivalent offline NJ tree from the local
# CI protein alignment, with tip labels matching data/CI_motif_states.csv.
treefile  <- here::here("data", "potyvirus_CI.treefile")
motif_csv <- here::here("data", "CI_motif_states.csv")

if (file.exists(treefile)) {
  ci_tree <- read.tree(treefile)

  # Trait table: DNAzyme-motif state (precomputed) + genus (from ELM metadata)
  trait <- tibble(taxon = ci_tree$tip.label)
  if (file.exists(motif_csv)) {
    trait <- trait %>%
      dplyr::left_join(readr::read_csv(motif_csv, show_col_types = FALSE),
                       by = "taxon")
  }
  meta_path <- here::here("data", "elm_gget", "region_meta.tsv")
  if (file.exists(meta_path)) {
    trait <- trait %>%
      dplyr::left_join(
        readr::read_tsv(meta_path, show_col_types = FALSE) %>%
          dplyr::select(taxon = name, genus),
        by = "taxon")
  }

  ggtree(ci_tree, layout = "circular", size = 0.25) %<+% trait +
    geom_tippoint(aes(colour = genus), size = 1.4, na.rm = TRUE) +
    ggplot2::scale_colour_viridis_d(option = "turbo", na.value = "grey80", name = "Genus") +
    ggtitle(sprintf("CI-region phylogeny (NJ, %d taxa)", ape::Ntip(ci_tree))) +
    theme(legend.position = "right", legend.key.size = unit(8, "pt"),
          legend.text = element_text(size = 6))
} else {
  cat("Tree file not found. Run scripts/build_ci_tree.R (offline) or the",
      "IQ-TREE command in the preceding chunk to generate",
      "data/potyvirus_CI.treefile.")
}
```

# Visualisation: Motif Sequence Logo

```{r logo}
#| eval: false
#| fig-width: 8
#| fig-height: 3
seqs_with_motif <- lapply(names(codon_CI), function(id) {
  pos    <- motif_pos_nt[id]
  if (is.na(pos)) return(NULL)
  seq_nt <- as.character(codon_CI[[id]])
  as.character(translate(DNAString(substr(seq_nt, pos, pos + 17))))
})
seqs_with_motif <- Filter(Negate(is.null), seqs_with_motif)
if (length(seqs_with_motif) > 0) {
  ggseqlogo(unlist(seqs_with_motif), method = "prob")
} else {
  cat("No motifs found for logo plot.")
}
```

# Test for Convergent Evolution

```{r tests}
#| eval: false
# Requires the codon motif states (motif_df, from the eval:false pipeline above)
# plus geiger::fitDiscrete and caper::phylo.d. Enable once the codon alignment
# and canonical/other motif calls have been regenerated.
trait      <- setNames(as.numeric(motif_df$motif == "canonical"), motif_df$taxon)
trait      <- trait[ci_tree$tip.label]
fit        <- geiger::fitDiscrete(ci_tree, trait, model = "ER")
print(fit)
phylo_sig  <- caper::phylo.d(data.frame(species = names(trait), x = trait), ci_tree, permut = 100)
print(phylo_sig)
```

# Summary

```{r summary}
#| eval: false
summary_tbl <- tibble(
  step  = c("Protein entries", "Reference CDS recovered", "Isolate CDS", "Codon alignment length (nt)"),
  value = c(length(prot_aln), sum(ref_diag$status == "ok"), length(cds_orf), width(codon_aln)[1])
)
print(summary_tbl)
```

# ICTV alignment — protein-to-codon pipeline

Starting from the protein alignment of Potyviridae used to generate the tree on
ICTV [@inouenagata2022], I added the amino acid sequences corresponding to the longest ORFs of the
initial DNAzyme blast hits (filtered to sequences >3000 amino acids, leaving 11 of
the original 16 hits).

```{r ictv-input}
ictv_aa_aln <- Biostrings::readAAStringSet(here::here("data", "OPSR.Poty.Fig3.v16_align.txt"))

aa_names <- tibble(seqname = names(ictv_aa_aln)) %>%
  tidyr::separate(seqname, into = c("seqname", "acc1", "acc2"), sep = "_", remove = FALSE)

ictv_tb <- readxl::read_xlsx(here::here("data", "VMR_MSL41.v1.20260320.xlsx"),
                              sheet = 2, col_types = "text") %>%
  dplyr::filter(Order == "Patatavirales") %>%
  tidyr::separate_rows(`Virus GENBANK accession`, sep = ";\\s*") %>%
  dplyr::mutate(genbank = stringr::str_remove(`Virus GENBANK accession`, "^.*:\\s*"))
ictv_tb
```

## Rolling GC content along the alignment

```{r rolling-gc-setup}
aln <- as.matrix(readDNAStringSet(here::here("data", "claude_polyproteinDNA_mafft.fasta")))
```

```{r rolling-gc-functions}
rolling_alignment_stats <- function(aln, window = 300, step = 30,
                                    codon_aligned = TRUE, remove_gaps = FALSE) {
  if (!is.matrix(aln))
    stop("Alignment must be a character matrix")
  if (codon_aligned && (window %% 3 != 0 || step %% 3 != 0))
    stop("window and step must be multiples of 3")
  n_sites    <- ncol(aln)
  nt         <- c("A", "C", "G", "T", "U")
  gc_content <- function(x) mean(x %in% c("G", "C"), na.rm = TRUE)
  shannon_entropy <- function(x) {
    x <- x[x %in% nt]
    n <- length(x); if (!n) return(NA_real_)
    cnt <- tabulate(match(x, nt), nbins = length(nt))   # fast vs table()
    p <- cnt[cnt > 0] / n
    -sum(p * log2(p))
  }
  cpg_oe <- function(x) {
    x <- x[x %in% c("A","C","G","T","U")]
    if (length(x) < 2) return(NA_real_)
    obs <- sum(x[-length(x)] == "C" & x[-1] == "G")
    exp <- sum(x == "C") * sum(x == "G") / length(x)
    if (exp == 0) return(NA_real_)
    obs / exp
  }
  results <- list(); win_id <- 1
  for (start in seq(1, n_sites - window + 1, by = step)) {
    end     <- start + window - 1
    sub_aln <- aln[, start:end, drop = FALSE]
    vec     <- as.vector(sub_aln)
    if (remove_gaps) vec <- vec[vec != "-"]
    codon_pos <- rep(1:3, length.out = window)
    pos1 <- vec[codon_pos == 1]; pos2 <- vec[codon_pos == 2]; pos3 <- vec[codon_pos == 3]
    results[[win_id]] <- data.frame(
      window = win_id, start = start, end = end,
      GC = gc_content(vec), GC3 = gc_content(pos3),
      CpG_OE = cpg_oe(vec), entropy = shannon_entropy(vec),
      gap_fraction = mean(vec == "-"),
      GC_pos1 = gc_content(pos1), GC_pos2 = gc_content(pos2), GC_pos3 = gc_content(pos3))
    win_id <- win_id + 1
  }
  do.call(rbind, results)
}
```

```{r compute-rolling-gc}
ras        <- rolling_alignment_stats(aln)
ras_degap  <- rolling_alignment_stats(aln, remove_gaps = TRUE)
```

```{r plot-rolling-entropy}
#| fig-cap: "Sliding-window Shannon entropy along the CI alignment. Low-entropy troughs mark the most conserved segments (functional constraint); peaks mark variable regions."
ras %>% ggplot2::ggplot(ggplot2::aes(x = start, y = entropy)) +
  ggplot2::geom_line() +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "Alignment position", y = "Shannon entropy")
```

```{r rolling-gc-per-sequence}
rolling_alignment_stats_per_sequence <- function(aln, window = 300, step = 30,
                                                  codon_aligned = TRUE, remove_gaps = FALSE) {
  if (!is.matrix(aln)) stop("Alignment must be a character matrix")
  if (codon_aligned && (window %% 3 != 0 || step %% 3 != 0))
    stop("window and step must be multiples of 3")
  n_sites    <- ncol(aln)
  nt         <- c("A", "C", "G", "T", "U")
  gc_content <- function(x) mean(x %in% c("G", "C"), na.rm = TRUE)
  shannon_entropy <- function(x) {
    x <- x[x %in% nt]
    n <- length(x); if (!n) return(NA_real_)
    cnt <- tabulate(match(x, nt), nbins = length(nt))   # fast vs table()
    p <- cnt[cnt > 0] / n
    -sum(p * log2(p))
  }
  cpg_oe <- function(x) {
    x <- x[x %in% nt]
    if (length(x) < 2) return(NA_real_)
    obs <- sum(x[-length(x)] == "C" & x[-1] == "G")
    exp <- sum(x == "C") * sum(x == "G") / length(x)
    if (exp == 0) return(NA_real_)
    obs / exp
  }
  # Preallocate per-window vectors and build one data frame per sequence,
  # binding once at the end (avoids per-window data.frame() + O(n^2) rbind).
  starts <- seq(1, n_sites - window + 1, by = step)
  nwin   <- length(starts)
  results <- vector("list", nrow(aln))
  for (i in seq_len(nrow(aln))) {
    seq_i <- aln[i, ]
    GC <- GC3 <- CpG <- ent <- gapf <- GC1 <- GC2 <- numeric(nwin)
    for (w in seq_len(nwin)) {
      st <- starts[w]; en <- st + window - 1
      ws <- seq_i[st:en]
      gapf[w] <- mean(ws == "-")
      wsc <- if (remove_gaps) ws[ws != "-"] else ws
      codon_pos <- rep(1:3, length.out = length(wsc))
      GC[w]  <- gc_content(wsc)
      GC3[w] <- gc_content(wsc[codon_pos == 3])
      CpG[w] <- cpg_oe(wsc)
      ent[w] <- shannon_entropy(wsc)
      GC1[w] <- gc_content(wsc[codon_pos == 1])
      GC2[w] <- gc_content(wsc[codon_pos == 2])
    }
    results[[i]] <- data.frame(
      sequence = rownames(aln)[i], window = seq_len(nwin),
      start = starts, end = starts + window - 1,
      GC = GC, GC3 = GC3, CpG_OE = CpG, entropy = ent, gap_fraction = gapf,
      GC_pos1 = GC1, GC_pos2 = GC2, GC_pos3 = GC3, stringsAsFactors = FALSE)
  }
  do.call(rbind, results)
}

ras_ps       <- rolling_alignment_stats_per_sequence(aln)
ras_ps_degap <- rolling_alignment_stats_per_sequence(aln, remove_gaps = TRUE)
```

Per-isolate sliding-window composition across the CI alignment: GC content (raw
and with alignment gaps removed) and the CpG observed/expected ratio. Tight
bundling of the lines means the compositional profile is conserved across
isolates; CpG O/E below the dashed line (= 1) indicates CpG suppression, a
signature linked to evasion of host nucleic-acid sensing.

```{r plot-gc-per-seq}
#| fig-cap:
#|   - "Per-isolate GC content along the CI alignment."
#|   - "Per-isolate GC content with alignment gaps removed."
#|   - "Per-isolate CpG observed/expected ratio (dashed line = 1, no suppression)."
library(ggplot2)
ggplot(ras_ps, aes(start, GC, colour = sequence)) +
  geom_line(alpha = 0.2) +
  theme_bw() +
  labs(x = "Genome position", y = "GC content") +
  theme(legend.position = "none")

ggplot(ras_ps_degap, aes(start, GC, colour = sequence)) +
  geom_line(alpha = 0.2) +
  theme_bw() +
  labs(x = "Genome position", y = "GC content (gaps removed)") +
  theme(legend.position = "none")

ggplot(ras_ps, aes(start, CpG_OE, colour = sequence)) +
  geom_line(alpha = 0.2) +
  geom_hline(yintercept = 1, linetype = 2) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "Genome position", y = "CpG O/E")
```
