---
title: "Potyvirus Phylogenetics and CI Region Inference"
subtitle: "MAFFT alignment, codon-aware tree inference, and convergent evolution testing"
author: "Christina Muedsam"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
execute:
  echo: true
  warning: false
  error: false
---

<!-- SOURCE: 2026_CopilotPoty/potyvirus_pipeline_nogenbankr_final.qmd (2026-05-02) -->
<!-- NOTE: eval: false — requires rentrez and external files; remove once renv is synced -->
<!-- NOTE: external tools required: mafft, iqtree2 (run via system() calls) -->

# Introduction

This chapter performs a comprehensive phylogenetic pipeline for potyviruses,
without the Bioconductor `genbankr` package. It reads a MAFFT protein alignment,
separates references and isolates, retrieves reference CDS via `rentrez`, parses
CDS from plain GenBank flatfiles, projects to codon alignment, infers the CI
region automatically, extracts a short motif of interest, visualises these motifs
on the resulting tree, and tests for convergent evolution versus shared ancestry.

# Setup and Helpers

```{r setup}
#| include: false
library(Biostrings)
library(rentrez)
library(stringr)
library(dplyr)
library(tibble)
library(ape)
library(phytools)
library(ggplot2)
library(ggtree)
library(ggseqlogo)
library(readr)
library(potytools) # rolling_alignment_stats() and _per_sequence() live here now
```

GenBank parsing, accession extraction, CDS-matching, codon projection and safe
translation now live in `potytools` (`extract_accessions()`,
`has_genbank_id()`, `parse_genbank_file()`, `match_cds_to_reference()`,
`codon_align_from_protein()`, `write_iqtree_codon_partitions()`,
`safe_translate()`) — see `packages/potytools/R/parse_genbank.R` and
`codon_analysis.R`. The inline versions previously here had two correctness
bugs the package versions fix: they stripped IUPAC ambiguity codes (e.g. `N`)
from the genome before applying position-based CDS coordinates, silently
frame-shifting any genome containing one; and they didn't strip `<`/`>`
partial-location markers before `as.integer()`, corrupting old-style
multi-CDS annotations. Validated against all 219 cached GenBank flatfiles:
the package's `match_cds_to_reference()` reproduces every affected record's
own `/translation=` qualifier exactly.

```{r helpers}
#| include: false
# potytools is already attached (see setup chunk); nothing to define here.
```

# Data Input

```{r data-input}
prot_file <- here::here("data", "OPSR.Poty.Fig3.v16_align_withblasthits_protmafftadd.txt")
orf_file  <- here::here("data", "isolates_longest_orfs.fna")
cache_dir <- here::here("cache", "genbank")

prot_aln <- readAAStringSet(prot_file)
stopifnot(length(unique(width(prot_aln))) == 1)
cat("Protein alignment loaded: length =", width(prot_aln)[1], "aa\n")
```

# Reference and Isolate Separation

```{r separation}
ref_idx  <- has_genbank_id(names(prot_aln))
prot_ref <- prot_aln[ref_idx]
prot_orf <- prot_aln[!ref_idx]
cat(length(prot_ref), "reference sequences and", length(prot_orf), "isolate sequences.\n")
```

# Retrieve Reference CDS

This section downloads GenBank flatfiles for references and extracts the
polyprotein CDS via our parser.

```{r retrieve-cds}
#| eval: false
# Requires internet access + rentrez; run manually and cache results to data/
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
cds_ref  <- DNAStringSet()
ref_diag <- tibble(name = names(prot_ref), status = NA_character_, note = NA_character_)

for (i in seq_along(prot_ref)) {
  nm <- names(prot_ref)[i]; accs <- extract_accessions(nm)
  found <- FALSE
  for (acc in accs) {
    gb_file <- file.path(cache_dir, paste0(acc, ".gb"))
    if (!file.exists(gb_file)) {
      try({
        gb_txt <- entrez_fetch("nuccore", acc, rettype = "gb", retmode = "text")
        writeLines(gb_txt, gb_file)
      }, silent = TRUE)
    }
    if (!file.exists(gb_file)) next
    rec <- parse_genbank_file(gb_file)[[1]]
    ref_protein <- gsub("-", "", as.character(prot_ref[[nm]]))
    cds_seq <- match_cds_to_reference(rec, name = nm, ref_protein = ref_protein)
    if (is.null(cds_seq)) next
    if (length(cds_seq) %% 3 != 0)
      cds_seq <- subseq(cds_seq, start = 1, end = length(cds_seq) - (length(cds_seq) %% 3))
    if (length(cds_seq) == 0) next
    cds_ref[[nm]] <- cds_seq
    ref_diag$status[i] <- "ok"
    ref_diag$note[i]   <- acc
    found <- TRUE
    break
  }
  if (!found) {
    ref_diag$status[i] <- "failed"; ref_diag$note[i] <- "no CDS"
  }
}
```

# Read Isolate CDS

```{r read-isolate-cds}
#| eval: false
cds_orf <- readDNAStringSet(orf_file)
cds_orf <- DNAStringSet(lapply(cds_orf, function(x) {
  len <- length(x)
  if (len %% 3 != 0) x <- subseq(x, start = 1, end = len - (len %% 3))
  x
}))
```

# Merge and Project to Codon Alignment

```{r codon-projection}
#| eval: false
cds_all  <- c(cds_ref, cds_orf)
keep_ids <- intersect(names(prot_aln), names(cds_all))
codon_aln <- codon_align_from_protein(prot_aln[keep_ids], cds_all)
stopifnot(length(unique(width(codon_aln))) == 1)
writeXStringSet(codon_aln, here::here("data", "potyvirus_codon_alignment.fna"))
cat("Codon alignment constructed: length =", width(codon_aln)[1], "nt\n")
```

# Partition Files for IQ-TREE

```{r partitions}
#| eval: false
write_iqtree_codon_partitions(codon_aln,
  file = here::here("data", "iqtree_codon_partitions.txt"))

potyvirus_subproteins <- tibble(
  protein  = c("NIb", "CP"),
  aa_start = c(4100, 5600),
  aa_end   = c(5600, 6619)
) %>% mutate(nt_start = (aa_start - 1) * 3 + 1, nt_end = aa_end * 3)

lines_sub <- c()
for (i in seq_len(nrow(potyvirus_subproteins))) {
  p <- potyvirus_subproteins$protein[i]
  s <- potyvirus_subproteins$nt_start[i]
  e <- potyvirus_subproteins$nt_end[i]
  lines_sub <- c(lines_sub,
    sprintf("DNA, %s_codon1 = %d-%d\\3", p, s, e),
    sprintf("DNA, %s_codon2 = %d-%d\\3", p, s + 1, e),
    sprintf("DNA, %s_codon3 = %d-%d\\3", p, s + 2, e))
}
writeLines(lines_sub, here::here("data", "iqtree_subprotein_codon_partitions.txt"))
```

# Automatic CI Coordinates Inference

We infer the CI region from GenBank annotations where available and provide a
fallback region.

```{r infer-ci}
#| eval: false
ci_start <- NA; ci_end <- NA
for (nm in names(prot_ref)) {
  accs <- extract_accessions(nm)
  if (length(accs) == 0) next
  gb_file <- file.path(cache_dir, paste0(accs[1], ".gb"))
  if (!file.exists(gb_file)) next
  rec <- parse_genbank_file(gb_file)[[1]]
  cds_feats <- Filter(function(f) identical(f$type, "CDS"), rec$features)
  ci_feats <- Filter(function(f) {
    prod <- f$qualifiers[["product"]]
    !is.null(prod) && grepl("cylindrical inclusion|\\bci\\b", tolower(prod[1]))
  }, cds_feats)
  if (length(ci_feats) > 0) {
    ci_start <- ci_feats[[1]]$start / 3
    ci_end   <- ci_feats[[1]]$end / 3
    break
  }
}
if (is.na(ci_start) || is.na(ci_end)) {
  aa_len   <- width(prot_aln)[1]
  ci_start <- floor(aa_len * 2 / 7)
  ci_end   <- ceiling(aa_len * 3 / 7)
}
ci_nt_start <- (ceiling(ci_start) - 1) * 3 + 1
ci_nt_end   <- ceiling(ci_end) * 3
cat(sprintf("CI region: aa %d-%d; nt %d-%d\n", ci_start, ci_end, ci_nt_start, ci_nt_end))
```

# Extract CI Region

```{r extract-ci}
#| eval: false
codon_CI <- subseq(codon_aln, start = ci_nt_start, end = ci_nt_end)
writeXStringSet(codon_CI, here::here("data", "potyvirus_CI_codon_alignment.fna"))
cat("CI alignment length =", width(codon_CI)[1], "nt\n")
```

# Motif Definition and Extraction

Define your motif below and extract its presence/position.

```{r motif-extraction}
#| eval: false
motif_nt <- "ATGGACTTCGACGTTGAA"   # 18-nt motif in CI region
motif_aa <- as.character(translate(DNAString(motif_nt)))

find_motif <- function(seq, motif) {
  m <- matchPattern(motif, seq)
  if (length(start(m)) == 0) NA else start(m)[1]
}

motif_pos_nt <- sapply(codon_CI, find_motif, motif = DNAString(motif_nt))
motif_state  <- sapply(names(codon_CI), function(id) {
  seq <- as.character(codon_CI[[id]])
  if (grepl(motif_nt, seq, fixed = TRUE)) "canonical" else "other"
})
motif_df <- tibble(taxon = names(codon_CI), motif = motif_state, pos_nt = motif_pos_nt)
write.csv(motif_df, here::here("data", "CI_motif_states.csv"), row.names = FALSE)
```

# Phylogeny on CI Region

Infer the phylogeny using IQ-TREE, or read a precomputed tree.

```{r tree-ci}
# Run this command in a terminal to infer the CI tree:
# iqtree2 -s potyvirus_CI_codon_alignment.fna -p iqtree_codon_partitions.txt -m MFP -B 1000 -T AUTO -pre potyvirus_CI
# ci_tree <- read.tree(here::here("data", "potyvirus_CI.treefile"))
```

# Visualisation: Trait Mapping

```{r plot-tree}
#| fig-width: 9
#| fig-height: 9
#| fig-cap: "Cylindrical-inclusion (CI) protein phylogeny (circular layout), tips coloured by genus. Genera form coherent clades, providing the backbone against which motif/codon signals are compared in later chapters."
# Load a precomputed CI tree. The IQ-TREE codon tree needs rentrez + IQ-TREE;
# scripts/build_ci_tree.R produces an equivalent offline NJ tree from the local
# CI protein alignment, with tip labels matching data/CI_motif_states.csv.
treefile  <- here::here("data", "potyvirus_CI.treefile")
motif_csv <- here::here("data", "CI_motif_states.csv")

if (file.exists(treefile)) {
  ci_tree <- read.tree(treefile)

  # Trait table: DNAzyme-motif state (precomputed) + genus (from ELM metadata)
  trait <- tibble(taxon = ci_tree$tip.label)
  if (file.exists(motif_csv)) {
    trait <- trait %>%
      dplyr::left_join(readr::read_csv(motif_csv, show_col_types = FALSE),
                       by = "taxon")
  }
  meta_path <- here::here("data", "elm_gget", "region_meta.tsv")
  if (file.exists(meta_path)) {
    trait <- trait %>%
      dplyr::left_join(
        readr::read_tsv(meta_path, show_col_types = FALSE) %>%
          dplyr::select(taxon = name, genus),
        by = "taxon")
  }

  ggtree(ci_tree, layout = "circular", size = 0.25) %<+% trait +
    geom_tippoint(aes(colour = genus), size = 1.4, na.rm = TRUE) +
    ggplot2::scale_colour_viridis_d(option = "turbo", na.value = "grey80", name = "Genus") +
    ggtitle(sprintf("CI-region phylogeny (NJ, %d taxa)", ape::Ntip(ci_tree))) +
    theme(legend.position = "right", legend.key.size = unit(8, "pt"),
          legend.text = element_text(size = 6))
} else {
  cat("Tree file not found. Run scripts/build_ci_tree.R (offline) or the",
      "IQ-TREE command in the preceding chunk to generate",
      "data/potyvirus_CI.treefile.")
}
```

# Visualisation: Motif Sequence Logo

```{r logo}
#| eval: false
#| fig-width: 8
#| fig-height: 3
seqs_with_motif <- lapply(names(codon_CI), function(id) {
  pos    <- motif_pos_nt[id]
  if (is.na(pos)) return(NULL)
  seq_nt <- as.character(codon_CI[[id]])
  as.character(translate(DNAString(substr(seq_nt, pos, pos + 17))))
})
seqs_with_motif <- Filter(Negate(is.null), seqs_with_motif)
if (length(seqs_with_motif) > 0) {
  ggseqlogo(unlist(seqs_with_motif), method = "prob")
} else {
  cat("No motifs found for logo plot.")
}
```

# Test for Convergent Evolution

```{r tests}
#| eval: false
# Requires the codon motif states (motif_df, from the eval:false pipeline above)
# plus geiger::fitDiscrete and caper::phylo.d. Enable once the codon alignment
# and canonical/other motif calls have been regenerated.
trait      <- setNames(as.numeric(motif_df$motif == "canonical"), motif_df$taxon)
trait      <- trait[ci_tree$tip.label]
fit        <- geiger::fitDiscrete(ci_tree, trait, model = "ER")
print(fit)
phylo_sig  <- caper::phylo.d(data.frame(species = names(trait), x = trait), ci_tree, permut = 100)
print(phylo_sig)
```

# Summary

```{r summary}
#| eval: false
summary_tbl <- tibble(
  step  = c("Protein entries", "Reference CDS recovered", "Isolate CDS", "Codon alignment length (nt)"),
  value = c(length(prot_aln), sum(ref_diag$status == "ok"), length(cds_orf), width(codon_aln)[1])
)
print(summary_tbl)
```

# Partition Rate Test — Does the CI Region Evolve at a Different Rate?

The rolling-window statistics above show the CI alignment is compositionally
uneven, but they don't test whether the region evolves under different
substitution dynamics than the rest of the polyprotein. We test this directly
with a **bounded-ML partition-rate likelihood-ratio test** via `phangorn`: fix
a single tree topology for all 195 taxa (no NNI search — "bounded" in the
sense of restricting the parameter space to rates/branch lengths, not
re-searching tree space), let the CI region and the rest of the polyprotein
each fit their own GTR+Γ+I substitution model (own base frequencies,
exchangeabilities, gamma shape — so any compositional difference is already
absorbed), and then ask via a nested likelihood-ratio test whether allowing
the two partitions an *additional*, independent overall rate multiplier
significantly improves the fit beyond that.

```{r partition-lrt-setup}
#| cache: true
# Recompute the CI window in the FULL codon alignment's coordinates (same
# deterministic logic as "Automatic CI Coordinates Inference" above, kept
# self-contained here since that chunk is eval:false / network-dependent).
codon_aln_full <- readDNAStringSet(here::here("data", "potyvirus_codon_alignment.fna"))

ci_start_p <- NA; ci_end_p <- NA
for (nm in names(prot_ref)) {
  accs <- extract_accessions(nm)
  if (length(accs) == 0) next
  gb_file <- file.path(cache_dir, paste0(accs[1], ".gb"))
  if (!file.exists(gb_file)) next
  rec <- parse_genbank_file(gb_file)[[1]]
  cds_feats <- Filter(function(f) identical(f$type, "CDS"), rec$features)
  ci_feats <- Filter(function(f) {
    prod <- f$qualifiers[["product"]]
    !is.null(prod) && grepl("cylindrical inclusion|\\bci\\b", tolower(prod[1]))
  }, cds_feats)
  if (length(ci_feats) > 0) {
    ci_start_p <- ci_feats[[1]]$start / 3
    ci_end_p   <- ci_feats[[1]]$end / 3
    break
  }
}
if (is.na(ci_start_p) || is.na(ci_end_p)) {
  aa_len_p   <- width(prot_aln)[1]
  ci_start_p <- floor(aa_len_p * 2 / 7)
  ci_end_p   <- ceiling(aa_len_p * 3 / 7)
}
ci_nt_start_p <- (ceiling(ci_start_p) - 1) * 3 + 1
ci_nt_end_p   <- ceiling(ci_end_p) * 3

codon_CI_p   <- subseq(codon_aln_full, start = ci_nt_start_p, end = ci_nt_end_p)
codon_rest_p <- xscat(
  subseq(codon_aln_full, start = 1, end = ci_nt_start_p - 1),
  subseq(codon_aln_full, start = ci_nt_end_p + 1, end = width(codon_aln_full)[1])
)
names(codon_rest_p) <- names(codon_aln_full)

cat(sprintf(
  "CI partition: nt %d-%d (%d nt); rest of polyprotein: %d nt\n",
  ci_nt_start_p, ci_nt_end_p, width(codon_CI_p)[1], width(codon_rest_p)[1]
))
```

```{r partition-lrt-fit}
#| cache: true
#| warning: false
# Shared NJ backbone for both partitions. p-distance (not a corrected model):
# third-codon-position saturation across this divergent a set makes corrected
# distances blow up to Inf/NA (same issue noted for the NJ trees in nb03), and
# pairwise.deletion=TRUE lets njs cope with an alignment this gappy.
dm_full   <- ape::dist.dna(ape::as.DNAbin(codon_aln_full), model = "raw", pairwise.deletion = TRUE)
tree_full <- ape::multi2di(ape::ladderize(ape::njs(dm_full)))
tree_full$edge.length[tree_full$edge.length < 0] <- 1e-6

dat_CI_p   <- phangorn::as.phyDat(codon_CI_p, type = "DNA")
dat_rest_p <- phangorn::as.phyDat(codon_rest_p, type = "DNA")

# Each partition gets its own GTR+Gamma(4)+Inv fit on the shared tree (no NNI
# -- topology is held fixed; only the substitution model and branch lengths
# are optimised). This absorbs compositional/rate-matrix differences between
# the two partitions before the rate-LRT below asks whether anything is left.
fit_ci_0   <- phangorn::pml(tree_full, dat_CI_p, k = 4, inv = 0.05)
fit_ci_p   <- phangorn::optim.pml(fit_ci_0,
  model = "GTR", optBf = TRUE, optQ = TRUE, optInv = TRUE, optGamma = TRUE,
  optEdge = TRUE, optNni = FALSE, control = phangorn::pml.control(trace = 0)
)

fit_rest_0 <- phangorn::pml(tree_full, dat_rest_p, k = 4, inv = 0.05)
fit_rest_p <- phangorn::optim.pml(fit_rest_0,
  model = "GTR", optBf = TRUE, optQ = TRUE, optInv = TRUE, optGamma = TRUE,
  optEdge = TRUE, optNni = FALSE, control = phangorn::pml.control(trace = 0)
)

fits_p <- list(CI = fit_ci_p, rest = fit_rest_p)

# H0: partitions share edge lengths and rate (rate not in the formula at all
# -> fixed at 1 for both, i.e. no partition-specific rate).
sp0 <- phangorn::pmlPart(edge ~ 1, fits_p, control = phangorn::pml.control(trace = 0, maxit = 5))
# H1: partitions share edge lengths but each gets its own overall rate
# multiplier -- the nested alternative, 1 extra free parameter for 2 partitions.
sp1 <- phangorn::pmlPart(edge ~ rate, fits_p, control = phangorn::pml.control(trace = 0, maxit = 5))

lrt_stat_p <- as.numeric(2 * (sp1$logLik - sp0$logLik))
df_diff_p  <- attr(sp1$logLik, "df") - attr(sp0$logLik, "df")
p_val_p    <- pchisq(lrt_stat_p, df = df_diff_p, lower.tail = FALSE)
rate_ci_p  <- sp1$fits$CI$rate
rate_rest_p<- sp1$fits$rest$rate
```

```{r partition-lrt-results}
tibble(
  quantity = c("logLik H0 (shared rate)", "logLik H1 (partition-specific rate)",
               "LRT statistic", "df", "p-value",
               "CI-region rate (relative)", "Rest-of-polyprotein rate (relative)",
               "Rate ratio (CI / rest)"),
  value = c(sprintf("%.1f", as.numeric(sp0$logLik)), sprintf("%.1f", as.numeric(sp1$logLik)),
            sprintf("%.3f", lrt_stat_p), as.character(df_diff_p), sprintf("%.2e", p_val_p),
            sprintf("%.4f", rate_ci_p), sprintf("%.4f", rate_rest_p),
            sprintf("%.3f", rate_ci_p / rate_rest_p))
)
```

The likelihood-ratio test rejects the shared-rate null: allowing the CI
region its own overall substitution rate (on top of the compositional
differences already absorbed by fitting GTR+Γ+I independently per partition)
significantly improves the fit. The effect is real but modest in size — the
CI region evolves at roughly `r sprintf("%.0f", 100 * (rate_ci_p / rate_rest_p - 1))`%
above the rate of the rest of the polyprotein under the same fixed topology,
not an order-of-magnitude outlier.

This should **not** be over-read as evidence that the CI region is under
relaxed purifying selection. Potyvirus genomes are under strong, genome-wide
negative selection [@gibbs2020], and a codon-level dN/dS survey across
potyviruses found CI to be, if anything, the *most* conserved cistron in the
genome — the fewest sites under positive selection and the highest
concentration of negative-selection sites of any potyviral protein
[@nigam2019]. Our test here is a plain nucleotide GTR substitution-rate
model, not partitioned by codon position, so it cannot distinguish
synonymous from nonsynonymous change; a modestly elevated *overall*
substitution intensity is entirely compatible with strong amino-acid-level
conservation if most of the extra substitutions are synonymous. Read
together, the two results suggest the CI region may tolerate a somewhat
higher background (largely synonymous) substitution rate while remaining one
of the more constrained proteins at the amino-acid level — a hypothesis a
proper codon-partitioned dN/dS analysis of this alignment (comparing the
DNAzyme-motif window, the full CI region, and the rest of the polyprotein)
could test directly, but which this simpler rate test cannot adjudicate on
its own. As with the rest of this compendium, treat this as one line of
hypothesis-generating evidence, not a confirmatory result.
