---
title: "Codon Usage and DNAzyme-like Motif Architecture"
subtitle: "RSCU, ENC, GC content, wobble bias and dinucleotide analysis of the DNAzyme-matching locus"
author: "chrissi m."
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
execute:
  echo: false
  warning: false
  message: false
---

<!-- SOURCE: 2026_cleanPoty/enhanced_dashboardwphylo_extended.qmd (2026-05-12, newest) -->
<!-- NOTE: source() calls replaced with library(potytools)                             -->
<!-- NOTE: run_complete_analysis_optimized() may need to be renamed to                 -->
<!--       run_complete_analysis() if not exported from potytools                      -->

```{r setup}
#| include: false
library(Biostrings)
library(ggplot2)
library(plotly)
library(DT)
library(dplyr)
library(tidyr)
library(cowplot)
library(seqinr)
library(ggseqlogo)
library(ape)
library(ggtree)
library(potytools)

theme_set(theme_cowplot())
```

```{r load-data}
#| include: false
#| cache: true

"%ni%" <- Negate("%in%")

### 18nt motif (names = accessions)
motif_seqs <- readDNAStringSet(here::here("data", "ci2cds_mafft_DNAzyme.fasta"))
motif_seqs_99 <- subseq(motif_seqs, start = 1, width = 99)

motif_tb <- tibble(
  seqname = gsub("\\/.*", "", names(motif_seqs)),
  start.ci = stringr::str_extract(names(motif_seqs), "(.*\\/)(.*)(\\-.*)", group = 2),
  end.ci = as.numeric(stringr::str_extract(names(motif_seqs), "(.*\\/)(.*\\-)(.*)", group = 3)) - 5,
  seq = as.character(motif_seqs_99))

motif_seqs <- DNAStringSet(dplyr::pull(motif_tb, seq, name = seqname))
DNAzyme_seqs <- readDNAStringSet(here::here("data", "claude_motif_18nt.fasta"))
ci_seqs <- readDNAStringSet(here::here("data", "claude_ci2cdsaln.fasta"))

ref_seqs <- readDNAStringSet(here::here("data", "claude_polyproteinDNA.fasta"))
ref_seqs <- ref_seqs[names(motif_seqs)]

metadata <- readr::read_delim(here::here("data", "2026-04-22_metadata_table.txt"), delim = "\t")
host_classification <- create_host_classification_table(metadata$full_name[(match(names(motif_seqs), metadata$accession))])
host_codon_tables <- create_host_specific_codon_tables(host_classification)
names(host_codon_tables) <- gsub(" .*", "", names(host_codon_tables))
host_classification$isolate <- gsub(" .*", "", host_classification$isolate)
metadata$isolate <- metadata$accession

metadata <- left_join(metadata, host_classification, suffix = c("", "")) %>%
  dplyr::mutate(DNAzyme_hit = case_when(
    stringr::str_detect(motif_seq, pattern = "GGCTAGCTACAACGA") ~ "match",
    !stringr::str_detect(motif_seq, pattern = "GGCTAGCTACAACGA") ~ "no_match",
    TRUE ~ NA))
```

```{r batch-analysis}
#| include: false
#| cache: true

# NOTE: run_complete_analysis_optimized() is the function name used in the original
# dashboard. If potytools only exports run_complete_analysis(), use that instead.
results <- run_complete_analysis_optimized(motif_seqs, ref_seqs, host_codon_tables,
                                           output_prefix = "new_improved")

motif_char <- as.character(motif_seqs)
ref_char   <- as.character(ref_seqs)

rscu_motif <- results$motif_rscu
rscu_ref   <- results$ref_rscu
gc_motif   <- results$gc_motif
gc_ref     <- results$gc_ref
enc_motif  <- results$enc_motif[!is.na(results$enc_motif)]
enc_ref    <- results$enc_ref[!is.na(results$enc_ref)]
```

```{r batch-analysis-dnazyme}
#| include: false
#| cache: true

results_DNAzyme <- run_complete_analysis_optimized(DNAzyme_seqs, ci_seqs, host_codon_tables,
                                                    output_prefix = "new_improved")

DNAzyme_char <- as.character(DNAzyme_seqs)
ci_char      <- as.character(ci_seqs)

rscu_DNAzyme <- results_DNAzyme$DNAzyme_rscu
rscu_ci      <- results_DNAzyme$ci_rscu
gc_DNAzyme   <- results_DNAzyme$gc_DNAzyme
gc_ci        <- results_DNAzyme$gc_ci
enc_DNAzyme  <- results_DNAzyme$enc_DNAzyme[!is.na(results_DNAzyme$enc_DNAzyme)]
enc_ci       <- results_DNAzyme$enc_ci[!is.na(results_DNAzyme$enc_ci)]
```

## Overview {.tabset}

### Dataset Summary

```{r summary-stats}
n_isolates <- length(motif_seqs)
n_monocot  <- sum(host_classification$host_type == "monocot", na.rm = TRUE)
n_dicot    <- sum(host_classification$host_type == "dicot", na.rm = TRUE)
n_unknown  <- sum(is.na(host_classification$host_type))
n_motifnt  <- unique(width(motif_seqs))

summary_table <- data.frame(
  Metric = c("Total Isolates", "Monocot-infecting", "Dicot-infecting",
             "Unknown Host", "Motif Length", "Mean Polyprotein Length"),
  Value = c(n_isolates, n_monocot, n_dicot, n_unknown, n_motifnt,
            sprintf("%.0f nt", mean(nchar(ref_char)))))

datatable(summary_table, options = list(dom = "t", pageLength = 10), rownames = FALSE)
```

### Codon Bias Metrics

```{r metrics-display}
metrics_table <- data.frame(
  Metric = c("Mean ENC (Motif)", "Mean ENC (Reference)",
             "Mean GC (Motif)",  "Mean GC (Reference)", "Mean GC3 (Motif)"),
  Value = c(
    sprintf("%.2f ± %.2f", mean(enc_motif), sd(enc_motif)),
    sprintf("%.2f ± %.2f", mean(enc_ref),   sd(enc_ref)),
    sprintf("%.3f ± %.3f", mean(gc_motif[,"GC_overall"]), sd(gc_motif[,"GC_overall"])),
    sprintf("%.3f ± %.3f", mean(gc_ref[,"GC_overall"]),   sd(gc_ref[,"GC_overall"])),
    sprintf("%.3f", mean(gc_motif[,"GC3"]))),
  Interpretation = c("Lower = more biased", "Baseline codon bias",
                     "Overall GC content",  "Baseline GC content", "Wobble position GC"))

datatable(metrics_table, options = list(dom = "t", pageLength = 10), rownames = FALSE)
```

### RSCU Validation

```{r rscu-validation}
rscu_max    <- max(results$rscu_comparison$RSCU_motif, na.rm = TRUE)
rscu_min    <- min(results$rscu_comparison$RSCU_motif, na.rm = TRUE)
rscu_mean   <- mean(results$rscu_comparison$RSCU_motif, na.rm = TRUE)
rscu_issues <- results$rscu_comparison$RSCU_motif > 6.01

validation_df <- data.frame(
  Check = c("Min RSCU", "Max RSCU", "Mean RSCU",
            "Values > 6 (invalid)", "RSCU Calculation"),
  Result = c(
    sprintf("%.3f", rscu_min), sprintf("%.3f", rscu_max), sprintf("%.3f", rscu_mean),
    sprintf("%d", sum(rscu_issues, na.rm = TRUE)),
    ifelse(rscu_max <= 6.01, "✓ CORRECT", "✗ ISSUE DETECTED")),
  Status = c(
    ifelse(rscu_min >= 0, "✓ Valid", "✗ Invalid"),
    ifelse(rscu_max <= 6.01, "✓ Valid", "✗ Invalid"),
    "Info",
    ifelse(sum(rscu_issues, na.rm = TRUE) == 0, "✓ Valid", "✗ Invalid"),
    ifelse(rscu_max <= 6.01, "✓ OK", "✗ ERROR")))

datatable(validation_df, options = list(dom = "t"), rownames = FALSE) %>%
  formatStyle("Status",
    backgroundColor = styleEqual(
      c("✓ Valid", "✓ OK", "Info", "✗ Invalid", "✗ ERROR"),
      c("#ccffcc", "#ccffcc", "#ffffcc", "#ffcccc", "#ffcccc")))
```

## Sequence Logos {.tabset}

```{r create-logos}
#| cache: true
#| fig.height: 10
#| fig.width: 14

logos         <- create_comprehensive_logo_analysis(dna_sequences = motif_seqs, host_classification)
DNAzyme_logos <- create_comprehensive_logo_analysis(dna_sequences = DNAzyme_seqs, host_classification)
```

### Combined DNA & Protein

```{r display-combined-logo}
#| fig.height: 10
#| fig.width: 14
DNAzyme_logos$combined_logo
logos$combined_logo
```

### DNA Only

```{r dna-logo}
#| fig.height: 5
#| fig.width: 14
logos$dna_logo
```

### Protein Only

```{r protein-logo}
#| fig.height: 5
#| fig.width: 14
logos$ic_protein
DNAzyme_logos$ic_protein
```

### Conservation Plot

```{r conservation-plot}
#| fig.height: 6
#| fig.width: 12
plot_information_content(motif_seqs,   title = "Nucleotide Conservation Across and Around Motif")
plot_information_content(DNAzyme_seqs, title = "Nucleotide Conservation Across 18nt Motif")
```

### Logo Statistics

```{r logo-stats-table}
logo_stats <- logo_statistics(motif_seqs)

datatable(
  logo_stats %>% mutate(across(c(entropy, conservation, frequency), ~round(., 3))),
  options = list(pageLength = 18, scrollX = TRUE), rownames = FALSE) %>%
  formatStyle("conservation",
    background = styleColorBar(logo_stats$conservation, "lightgreen"),
    backgroundSize = "100% 90%", backgroundRepeat = "no-repeat",
    backgroundPosition = "center")
```

```{r DNAzyme-stats-table}
DNAzyme_stats <- logo_statistics(DNAzyme_seqs)

datatable(
  DNAzyme_stats %>% mutate(across(c(entropy, conservation, frequency), ~round(., 3))),
  options = list(pageLength = 18, scrollX = TRUE), rownames = FALSE) %>%
  formatStyle("conservation",
    background = styleColorBar(logo_stats$conservation, "lightgreen"),
    backgroundSize = "100% 90%", backgroundRepeat = "no-repeat",
    backgroundPosition = "center")
```

## Predicted DNAzyme Activity {.tabset}

The DNAzyme-like locus matches the catalytic core of the **10-23 RNA-cleaving
deoxyribozyme** (15 nt, `GGCTAGCTACAACGA`) [@santoro1997]. Whether a given
homolog would retain catalytic activity depends on conservation of this core:
it is highly intolerant to substitution [@zaborowska2002], and position **G14**
is the catalytic general base that activates the 2'-OH nucleophile
[@borggraefe2022; @cramer2023]. Below, each homolog's best-matching core window is
scored against the canonical core with per-position criticality weights (a
sequence-based heuristic, not a catalytic-rate measurement).

### Catalytic-core criticality

```{r dnazyme-core-table}
core_tbl <- potytools::dnazyme_core_table()
datatable(core_tbl, rownames = FALSE,
          options = list(pageLength = 15, dom = "t"),
          caption = "10-23 catalytic core: per-position criticality weights.") %>%
  formatStyle("weight",
    background = styleColorBar(c(0, 1), "salmon"),
    backgroundSize = "100% 90%", backgroundRepeat = "no-repeat",
    backgroundPosition = "center")
```

### Activity scores

```{r dnazyme-activity-score}
act <- potytools::score_dnazyme_activity(motif_seqs)

# attach host / DNAzyme-match context where available
act_ctx <- act %>%
  dplyr::left_join(
    metadata %>% dplyr::select(id = accession, host_type, DNAzyme_hit),
    by = "id")

datatable(
  act_ctx %>% dplyr::select(id, core_window, n_substitutions,
                            critical_substitutions, weighted_identity,
                            activity_score, predicted_active, host_type),
  rownames = FALSE, options = list(pageLength = 12, scrollX = TRUE),
  caption = "Predicted 10-23 activity per homolog (core conservation heuristic).") %>%
  formatStyle("activity_score",
    background = styleColorBar(c(0, 1), "lightgreen"),
    backgroundSize = "100% 90%", backgroundRepeat = "no-repeat",
    backgroundPosition = "center") %>%
  formatStyle("predicted_active",
    backgroundColor = styleEqual(c(TRUE, FALSE), c("#d5f5e3", "#fadbd8")))
```

### Score distribution

```{r dnazyme-activity-dist}
#| fig.height: 4
#| fig.width: 9
ggplot(act, aes(x = activity_score)) +
  geom_histogram(bins = 20, fill = "#2E86AB", alpha = 0.8, colour = "white") +
  labs(title = "Predicted DNAzyme activity across CI DNAzyme-like homologs",
       subtitle = "Multiplicative core-conservation model (1 = canonical core; 0 = essential-position substitution)",
       x = "Predicted activity score", y = "Number of homologs") +
  theme_cowplot()
```

### Core conservation logo

```{r dnazyme-core-conservation}
#| fig.height: 3.5
#| fig.width: 9
core_windows <- act$core_window[!is.na(act$core_window)]
if (length(core_windows) > 0) {
  ggseqlogo(core_windows, method = "prob") +
    ggtitle("Observed catalytic-core windows vs canonical GGCTAGCTACAACGA") +
    theme(plot.title = element_text(size = 11))
} else {
  cat("No core windows detected.")
}
```

## Phylogenetic Analysis {.tabset}

```{r phylogenetic-setup}
#| cache: true
#| include: false

# Neighbour-joining trees for each region with bootstrap support. Uses the raw
# p-distance (bounded, robust to the sequence saturation that makes corrected
# distances like TN93 diverge to Inf on the deeper CI/polyprotein alignments)
# and NJ (njs tolerates incomplete matrices). Requires phangorn; freeze caches
# the result so it is computed once.
run_phylo <- requireNamespace("phangorn", quietly = TRUE)
n_boot    <- 30

trees            <- NULL
tree_comparisons <- list()

if (run_phylo) {
  pp_dna <- readDNAStringSet(here::here("data", "claude_polyproteinDNA_mafft.fasta"))
  names(pp_dna) <- gsub(" .*", "", names(pp_dna))

  ci_cds <- readDNAStringSet(here::here("data", "claude_ci2cdsaln.fasta"))
  names(ci_cds) <- gsub(" .*", "", names(ci_cds))

  regions <- list(motif = motif_seqs, ci_cds = ci_cds,
                  pp_dna = pp_dna, DNAzyme = DNAzyme_seqs)

  trees <- list()
  for (region_name in names(regions)) {
    cat(sprintf("\n=== Building tree for %s ===\n", region_name))
    tree <- tryCatch(
      build_phylo_tree(regions[[region_name]],
                       method = "raw", tree_method = "nj", bootstrap = n_boot),
      error = function(e) { message("  skipped ", region_name, ": ", conditionMessage(e)); NULL })
    if (!is.null(tree)) trees[[region_name]] <- tree
  }
  if (length(trees) == 0) trees <- NULL
}
```

```{r tree-comparisons}
#| include: false
#| cache: true

if (!is.null(trees) && length(trees) > 0) {
  region_names <- names(trees)
  for (i in 1:(length(region_names) - 1)) {
    for (j in (i + 1):length(region_names)) {
      region1 <- region_names[i]; region2 <- region_names[j]
      comp    <- paste(region1, region2, sep = "_vs_")
      cat(sprintf("\nComparing %s:\n", comp))
      tree_comparisons[[comp]] <- compare_trees(trees[[region1]], trees[[region2]])
    }
  }
}
```

### Overview

```{r phylo-overview}
if (!is.null(trees)) {
  cat("Phylogenetic trees built for", length(trees), "genomic regions\n\n")
  for (region in names(trees)) cat(sprintf("  - %s\n", region))
  cat("\nTree comparisons performed:\n")
  for (comparison in names(tree_comparisons)) cat(sprintf("  - %s\n", comparison))
} else {
  cat("Phylogenetic analysis not available.\n")
}
```

### Tree Comparison

```{r phylo-comparison-table}
if (!is.null(trees) && length(tree_comparisons) > 0) {
  comp_summary <- data.frame()
  for (comp_name in names(tree_comparisons)) {
    comp <- tree_comparisons[[comp_name]]
    if (!is.null(comp)) {
      comp_summary <- rbind(comp_summary, data.frame(
        Comparison = comp_name, N_Taxa = comp$n_taxa,
        RF_Distance = round(comp$RF_distance, 3),
        Cophenetic_Correlation = round(comp$cophenetic_correlation, 3),
        Interpretation = ifelse(comp$RF_distance < 0.3, "Similar topology",
                         ifelse(comp$RF_distance < 0.7, "Moderately different",
                                "Very different"))))
    }
  }
  if (nrow(comp_summary) > 0) {
    datatable(comp_summary, options = list(pageLength = 10), rownames = FALSE) %>%
      formatStyle("RF_Distance",
        background = styleColorBar(comp_summary$RF_Distance, "lightcoral"),
        backgroundSize = "100% 90%", backgroundRepeat = "no-repeat",
        backgroundPosition = "center")
  }
} else {
  cat("No tree comparisons available")
}
```

### Phylogenetic Trees

```{r phylo-trees}
#| fig.height: 10
#| fig.width: 12
if (!is.null(trees)) {
  for (region_name in names(trees)) {
    p <- ggtree(trees[[region_name]]) +
      geom_tiplab(size = 2) +
      geom_nodelab(aes(label = round(as.numeric(label), 0)), size = 2, hjust = -0.3, color = "blue") +
      labs(title = paste("Phylogeny:", region_name)) +
      theme_tree2()
    print(p)
  }
} else {
  cat("Phylogenetic trees not available")
}
```

### Interpretation

```{r phylo-interpretation}
if (!is.null(trees) && length(tree_comparisons) > 0) {
  comp_rf_values <- sapply(tree_comparisons, function(x) if (!is.null(x)) x$RF_distance else NA)
  comp_rf_values <- comp_rf_values[!is.na(comp_rf_values)]

  if (length(comp_rf_values) > 0) {
    most_similar  <- names(which.min(comp_rf_values))
    most_different <- names(which.max(comp_rf_values))
    cat(sprintf("- **Most similar phylogenies**: %s (RF = %.3f)\n", most_similar, min(comp_rf_values)))
    cat(sprintf("- **Most different phylogenies**: %s (RF = %.3f)\n\n", most_different, max(comp_rf_values)))

    if (max(comp_rf_values) > 0.7) {
      cat("- Very different topologies suggest:\n")
      cat("  - **Recombination** between viral lineages\n")
      cat("  - **Different selection pressures** on different genes\n")
      cat("  - **Horizontal gene transfer** events\n")
    }

    motif_comps <- comp_rf_values[grepl("motif", names(comp_rf_values))]
    if (length(motif_comps) > 0 && mean(motif_comps) > 0.5) {
      cat("\n**18nt Motif shows distinct phylogenetic signal** - may indicate:\n")
      cat("- Functional constraint specific to this region\n")
      cat("- Recent recombination affecting other regions\n")
    }
  }
}
```

## Codon Usage {.tabset}

### motif vs. genome

```{r cu-comparison}
#| cache: true
cu_comparison <- compare_codon_usage_fast(motif_seqs, ref_seqs)
```

#### Codon usage

```{r cu-comparison-table}
datatable(
  cu_comparison %>%
    mutate(log2FC = log2(fold_change), nlt_padj = -log10(p_adjusted)) %>%
    select(codon, amino_acid, log2FC, nlt_padj) %>%
    mutate(across(where(is.numeric), ~round(., 3))),
  options = list(pageLength = 20, scrollX = TRUE), filter = "top", rownames = FALSE)
```

#### Top Differences (motif vs genomes)

```{r rscu-top-diffs}
#| fig.height: 6
#| fig.width: 10
top_diffs_cu <- cu_comparison %>% arrange(desc(abs(fold_change))) %>% head(20)

ggplot(top_diffs_cu, aes(x = reorder(codon, fold_change), y = fold_change,
                         fill = -log10(p_adjusted))) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "solid", color = "black") +
  labs(title = "Top 20 Codons with Largest CU Differences",
       subtitle = "Motif vs Reference",
       x = "Codon", y = "CU Fold Change (Motif vs Reference)") +
  coord_flip() + theme_cowplot() + theme(legend.position = "top")
```

#### Interactive CU Table

```{r rscu-table}
datatable(
  cu_comparison %>%
    select(codon, amino_acid, frequency_motif, frequency_ref, fold_change, p_adjusted,
           count_motif, count_ref) %>%
    mutate(across(where(is.numeric), ~round(., 3))),
  options = list(pageLength = 20, scrollX = TRUE), filter = "top", rownames = FALSE) %>%
  formatStyle("fold_change",
    background = styleColorBar(range(-log10(cu_comparison$p_adjusted)), "lightblue"),
    backgroundSize = "100% 90%", backgroundRepeat = "no-repeat",
    backgroundPosition = "center")
```

### DNAzyme-match vs other

```{r cu-comp-DNAzyme}
DNAzyme_iso <- metadata$isolate[(metadata$DNAzymeseq == "DNAzyme")]
DNAzyme_seq <- motif_seqs[(names(motif_seqs) %in% DNAzyme_iso)]

other_iso <- metadata$isolate[(metadata$DNAzymeseq == "other")]
other_seq <- motif_seqs[(names(motif_seqs) %in% other_iso)]

cu_comp_DNAzyme <- compare_codon_usage_fast(DNAzyme_seq, other_seq)
```

#### Top Differences

```{r cu-top-diffs-DNAzyme}
#| fig.height: 6
#| fig.width: 10
top_diffs_cu_DNAzyme <- cu_comp_DNAzyme %>% arrange(desc(abs(fold_change))) %>% head(20)

ggplot(top_diffs_cu_DNAzyme, aes(x = reorder(codon, fold_change), y = fold_change,
                                  fill = -log10(p_adjusted))) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "solid", color = "black") +
  labs(title = "Top 20 Codons with Largest CU Differences",
       subtitle = "DNAzyme-matches vs. other",
       x = "Codon", y = "CU Fold Change (DNAzyme-matches vs. other)") +
  coord_flip() + theme_cowplot() + theme(legend.position = "top")
```

## Host-Specific Analysis {.tabset}

```{r host-analysis}
cai_results <- calculate_host_specific_cai(motif_seqs, host_codon_tables)
```

### CAI Distribution

```{r cai-plot}
#| fig.height: 6
#| fig.width: 8
host_classification_byacc          <- host_classification
host_classification_byacc$full_name <- host_classification_byacc$isolate
host_classification_byacc$isolate   <- gsub(" .*", "", host_classification_byacc$isolate)

plot_cai_by_host(cai_results, host_classification_byacc)
```

### Host Comparison

```{r host-comparison}
#| fig.height: 10
#| fig.width: 14
groups_host <- group_sequences_by_host_type(motif_seqs, host_classification)
monocots <- paste(groups_host$monocot, collapse = "")
dicots   <- paste(groups_host$dicot,   collapse = "")

if (nchar(dicots) %% 3 == 0 && nchar(monocots) %% 3 == 0) {
  monocots_rscu <- calculate_rscu_correct(monocots) %>% dplyr::mutate(host_type = "monocot")
  dicots_rscu   <- calculate_rscu_correct(dicots)   %>% dplyr::mutate(host_type = "dicot")

  combined_rscu <- bind_rows(monocots_rscu, dicots_rscu)

  ggplot(combined_rscu, aes(x = reorder(codon, RSCU), y = RSCU, fill = host_type)) +
    geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
    labs(title = "RSCU: Monocot vs Dicot-infecting Potyviruses",
         x = "Codon", y = "RSCU") +
    coord_flip() + theme_cowplot() + theme(legend.position = "top")
}
```

## RSCU Analysis {.tabset}

```{r rscu-setup}
rscu_matrix <- calculate_rscu_matrix_fast(motif_seqs)
```

### RSCU Heatmap

```{r rscu-heatmap}
#| fig.height: 10
#| fig.width: 14
if (!is.null(rscu_matrix)) {
  heatmap_data <- as.data.frame(rscu_matrix) %>%
    tibble::rownames_to_column("codon") %>%
    tidyr::pivot_longer(-codon, names_to = "isolate", values_to = "RSCU")

  ggplot(heatmap_data, aes(x = isolate, y = codon, fill = RSCU)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 1) +
    labs(title = "RSCU Heatmap — Motif Sequences") +
    theme_minimal() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
}
```

## Multivariate Analysis {.tabset}

```{r pca-setup}
#| cache: true
pca_results <- perform_rscu_pca_improved(motif_seqs, host_classification)
```

The first principal components summarise how each isolate's relative synonymous
codon usage (RSCU) deviates from the mean. Isolates that cluster together share a
codon-usage signature; separation by host type would indicate host-driven codon
adaptation.

### PCA Biplot

```{r pca-biplot}
#| fig.height: 8
#| fig.width: 10
#| fig-cap: "PCA of RSCU codon usage (PC1 vs PC2), each point an isolate coloured by host type. Axis labels report the variance explained by each component."
if (!is.null(pca_results)) {
  plot_pca_biplot(pca_results, host_classification)
}
```

### PCA 3D (interactive)

```{r pca-3d}
#| fig.height: 8
#| fig.width: 10
#| fig-cap: "Interactive 3D RSCU-PCA (PC1–PC3). Drag to rotate; hover a point for the isolate accession. Adding the third component resolves structure hidden in the 2D projection."
if (!is.null(pca_results)) {
  plot_pca_3d(pca_results, host_classification)
}
```

## Advanced Metrics {.tabset}

### GC Content by Codon Position

```{r gc-by-position}
#| fig.height: 6
#| fig.width: 10
gc_combined <- bind_rows(
  as.data.frame(gc_motif) %>% dplyr::mutate(group = "motif"),
  as.data.frame(gc_ref)   %>% dplyr::mutate(group = "polyprotein"))

gc_long <- gc_combined %>%
  select(GC_pos1 = GC1, GC_pos2 = GC2, GC_pos3 = GC3, group) %>%
  tidyr::pivot_longer(-group, names_to = "position", values_to = "GC")

ggplot(gc_long, aes(x = position, y = GC, fill = group)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "GC Content by Codon Position",
       x = "Codon position", y = "GC fraction") +
  theme_cowplot()
```

### Dinucleotide Bias (CpG suppression)

RNA viruses of plants and animals are frequently depleted in the **CpG** (and
often **UpA**) dinucleotide relative to what their base composition predicts — a
signature associated with evasion of host nucleic-acid immune sensing
[@greenbaum2008; @takata2017]. The odds ratio $\rho = f_{XY} / (f_X f_Y)$
measures observed-vs-expected dinucleotide frequency: $\rho < 1$ indicates
suppression, $\rho > 1$ over-representation.

```{r dinucleotide-oe}
#| fig.height: 5
#| fig.width: 10
#| fig-cap: "Dinucleotide odds ratios (ρ = observed/expected) across the DNAzyme-matching motif locus. Bars below the dashed line (ρ = 1) are suppressed; CpG is highlighted. Marked CpG suppression is consistent with an immune-evasion signature."
dinuc <- calculate_dinucleotide_freq(motif_seqs)
dinuc$highlight <- ifelse(dinuc$dinucleotide == "CG", "CpG", "other")

ggplot(dinuc, aes(x = reorder(dinucleotide, rho), y = rho, fill = highlight)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = 1, linetype = 2) +
  scale_fill_manual(values = c(CpG = "#D62828", other = "#2E86AB"),
                    name = NULL) +
  labs(title = "Dinucleotide odds ratios across the motif locus",
       x = "Dinucleotide", y = expression(rho ~ "(observed / expected)")) +
  theme_cowplot() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Summary Report

```{r summary-report}
cat("## Potyvirus DNAzyme-like Motif Analysis — Summary\n\n")
cat(sprintf("- Sequences analysed: %d isolates\n", n_isolates))
cat(sprintf("- Motif length: %d nt\n", n_motifnt))
cat(sprintf("- Mean ENC (motif): %.2f\n", mean(enc_motif)))
cat(sprintf("- Mean ENC (polyprotein): %.2f\n", mean(enc_ref)))
cat(sprintf("- Mean GC3 (motif): %.3f\n", mean(gc_motif[,"GC3"])))
cat(sprintf("- DNAzyme exact matches: %d\n", sum(metadata$DNAzyme_hit == "match", na.rm = TRUE)))
```

## ICTV sequences

```{r ictv-mafft}
#| cache: true

ictv_aa_aln <- readAAStringSet(here::here("data", "OPSR.Poty.Fig3.v16_align.txt"))

ictv_tb <- readxl::read_xlsx(here::here("data", "VMR_MSL41.v1.20260320.xlsx"),
                              sheet = 2, col_types = "text") %>%
  dplyr::filter(Order == "Patatavirales") %>%
  tidyr::separate_rows(`Virus GENBANK accession`, sep = ";\\s*") %>%
  dplyr::mutate(genbank = stringr::str_remove(`Virus GENBANK accession`, "^.*:\\s*"))
```

```{r poty-best-polyprotein}
poty_best_polyprotein <- function(seqs, min_len = 2000) {
  by_acc <- split(seqs, gsub("\\..*", "", names(seqs)))
  best <- lapply(by_acc, function(x) x[which.max(width(x))])
  best <- do.call(c, best)
  best[width(best) >= min_len]
}
```

```{r genbank-reader}
#| cache: true

read_genbank_folder <- function(path, pattern = "\\.gb$") {
  files <- list.files(path, pattern = pattern, full.names = TRUE)
  seqs  <- DNAStringSet()
  for (f in files) {
    lines    <- readLines(f)
    acc_line <- grep("^ACCESSION", lines, value = TRUE)
    acc      <- if (length(acc_line)) gsub("ACCESSION +", "", acc_line[1]) else basename(f)
    origin   <- grep("^ORIGIN", lines)
    if (!length(origin)) next
    raw  <- lines[(origin + 1):length(lines)]
    raw  <- raw[!grepl("^//", raw)]
    seq  <- paste(gsub("[^acgtACGT]", "", raw), collapse = "")
    seqs <- c(seqs, setNames(DNAStringSet(seq), acc))
  }
  seqs
}
```

```{r codon-align}
codon_align_from_protein_local <- function(prot_aln, cds) {
  out <- DNAStringSet()
  for (id in intersect(names(prot_aln), names(cds))) {
    prot   <- strsplit(as.character(prot_aln[[id]]), "")[[1]]
    codons <- substring(as.character(cds[[id]]),
                        seq(1, length(cds[[id]]), 3),
                        seq(3, length(cds[[id]]), 3))
    idx <- 1; seq_out <- character()
    for (aa in prot) {
      if (aa == "-") {
        seq_out <- c(seq_out, "---")
      } else {
        seq_out <- c(seq_out, if (idx > length(codons)) "NNN" else codons[idx])
        idx <- idx + 1
      }
    }
    out[[id]] <- DNAString(paste(seq_out, collapse = ""))
  }
  out
}
```

## Rolling genome-wide composition {.tabset}

A sliding window along the polyprotein alignment maps how compositional bias
varies from region to region, rather than as a single genome-wide average. Local
dips in **GC3** (third-codon-position GC) track translational/codon adaptation,
troughs in **CpG O/E** localise immune-evasion signal to particular genes, and
**Shannon entropy** marks the most conserved (low-entropy) versus most variable
segments. Windows are computed with `potytools::rolling_alignment_stats()`
(300 bp window, 30 bp step, codon-aware).

```{r rolling-gc-read}
aln <- as.matrix(readDNAStringSet(here::here("data", "claude_polyproteinDNA_mafft.fasta")))
```

```{r rolling-gc-compute}
#| cache: true
ras    <- rolling_alignment_stats(aln)
ras_ps <- rolling_alignment_stats_per_sequence(aln)
```

### GC, GC3 & entropy

```{r rolling-gc-tracks}
#| fig.height: 8
#| fig.width: 11
#| fig-cap: "Sliding-window GC, third-position GC (GC3) and Shannon entropy across the polyprotein alignment (300 bp window, 30 bp step). Conserved, compositionally biased regions appear as coordinated GC3 shifts with low entropy."
ras_long <- ras %>%
  select(start, GC, GC3, entropy) %>%
  tidyr::pivot_longer(-start, names_to = "metric", values_to = "value")

ggplot(ras_long, aes(start, value)) +
  geom_line(colour = "#2E86AB") +
  facet_wrap(~metric, ncol = 1, scales = "free_y") +
  labs(x = "Alignment position (bp)", y = NULL) +
  theme_cowplot()
```

### CpG O/E

```{r rolling-cpg}
#| fig.height: 4
#| fig.width: 11
#| fig-cap: "Sliding-window CpG observed/expected ratio across the polyprotein. Values below the dashed line (O/E = 1) indicate regional CpG suppression."
ggplot(ras, aes(start, CpG_OE)) +
  geom_line(colour = "#D62828") +
  geom_hline(yintercept = 1, linetype = 2) +
  labs(x = "Alignment position (bp)", y = "CpG O/E") +
  theme_cowplot()
```

### Per-isolate GC3

```{r rolling-gc-per-seq}
#| fig.height: 5
#| fig.width: 11
#| fig-cap: "Per-isolate sliding-window GC3, one line per isolate. Tight bundling indicates the compositional profile is conserved across the lineage; local fanning marks isolate-specific variation."
ggplot(ras_ps, aes(start, GC3, group = sequence)) +
  geom_line(alpha = 0.25, colour = "#2E86AB") +
  labs(x = "Alignment position (bp)", y = "GC3") +
  theme_cowplot()
```
