---
title: "Protein Motif Analysis"
subtitle: "Eukaryotic Linear Motifs in Potyviridae CI helicase — gget/ELM search and genus-specificity"
author: "chrissi m."
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    page-layout: full
execute:
  echo: true
  warning: false
  message: false
  error: true
---

## Overview

The cylindrical inclusion (CI) helicase carries a short conserved region that,
across Potyviridae, overlaps a number of **eukaryotic linear motifs (ELMs)** —
short functional peptide motifs catalogued in the ELM database
[@elm2024]. Here I extract the CI motif-flanking region from each isolate in the
ICTV Potyviridae alignment [@inouenagata2022], search it against ELM, and ask which
motifs are **characteristic of particular genera**.

The ELM search is run with the [`gget`](https://github.com/pachterlab/gget)
Python package (`gget elm`), which reproduces the ELM webserver's two search
modes: an ortholog (DIAMOND) search against ELM-annotated proteins and a direct
regex search. It runs in a self-contained `pixi` environment inside WSL; the
per-isolate results are written to `data/elm_gget/` by
`scripts/prep_elm_regions.R` → `scripts/gget_elm_run.py` →
`scripts/viz_elm_motifs.R`. This chapter reads those results.

```{r setup}
#| include: false
library(Biostrings)
library(tidyverse)
library(DECIPHER)   # gap removal
library(potytools)

data_dir <- here::here("data")
elm_dir  <- here::here("data", "elm_gget")
```

## CI motif region

I anchor the CI region on the alignment consensus at the conserved `VASYN`
motif (allowing one mismatch) and take a ±20 aa window. The alignment is read
from the **local** ICTV file (`data/OPSR.Poty.Fig3.v16_align.txt`).

```{r region}
protaln <- readAAStringSet(file.path(data_dir, "OPSR.Poty.Fig3.v16_align.txt"),
                           format = "fasta")

cons    <- consensusString(protaln, ambiguityMap = "X")
m       <- matchPattern("VASYN", cons, max.mismatch = 1)
anchora <- start(m)[1]
anchorb <- end(m)[1]

region  <- subseq(protaln, anchora - 20, anchorb + 20)   # aligned window
cat(sprintf("VASYN anchor at consensus %d-%d; region columns %d-%d (%d isolates)\n",
            anchora, anchorb, anchora - 20, anchorb + 20, length(region)))
region
```

## Isolate metadata and genus assignment

Each sequence name encodes a virus abbreviation and one or more accessions.
Genus is assigned by joining to the ICTV VMR table [@ictvvmr2024] — first by GENBANK
accession, then, for isolates that are not VMR species exemplars, by matching
the virus abbreviation to the VMR *"Virus name abbreviation(s)"* column. This
two-stage join (implemented in `scripts/prep_elm_regions.R`) leaves **no**
isolate unclassified.

```{r meta}
meta <- readr::read_tsv(file.path(elm_dir, "region_meta.tsv"),
                        show_col_types = FALSE)

meta %>%
  count(genus, genus_source, name = "n_isolates") %>%
  arrange(desc(n_isolates)) %>%
  knitr::kable(caption = "Isolates per genus and how genus was assigned.")
```

## ELM search results (gget)

`gget elm` was run per isolate in ortholog + regex mode. As expected for viral
CI proteins, the **ortholog** search returns nothing (ELM's ortholog set is
eukaryotic), so the **regex** motif matches carry the signal — the same content
the ELM webserver returns for a direct sequence query.

```{r read-gget}
regex_path <- file.path(elm_dir, "elm_regex.tsv")
have_gget  <- file.exists(regex_path)

if (have_gget) {
  elm <- readr::read_tsv(regex_path, show_col_types = FALSE) %>%
    dplyr::left_join(dplyr::select(meta, query = name, virus_abbr, genus),
                     by = "query")
  cat(sprintf("%d motif hits across %d isolates and %d ELM motif classes\n",
              nrow(elm), dplyr::n_distinct(elm$query),
              dplyr::n_distinct(elm$ELMIdentifier)))
  elm %>%
    dplyr::count(ELMType, name = "hits") %>%
    dplyr::arrange(dplyr::desc(hits)) %>%
    knitr::kable(caption = "ELM hits by motif type.")
} else {
  message("data/elm_gget/elm_regex.tsv not found. ",
          "Regenerate with scripts/prep_elm_regions.R -> gget_elm_run.py.")
}
```

## Motif landscape across isolates

Motif presence per isolate, ELM classes on **y**, isolate fragments on **x**,
faceted and coloured by genus (each panel is one genus, width proportional to
isolate count). The wide bands are motifs shared across essentially all
*Potyvirus* isolates; the narrow genus panels reveal motifs restricted to
particular lineages.

```{r grid}
#| fig-width: 15
#| fig-height: 18
#| eval: !expr file.exists(file.path(here::here("data","elm_gget"),"elm_regex.tsv"))
pa <- elm %>% dplyr::distinct(query, genus, ELMIdentifier)

genus_sizes <- meta %>% dplyr::count(genus, name = "n_iso") %>%
  dplyr::arrange(dplyr::desc(n_iso))
isolate_order <- meta %>%
  dplyr::mutate(genus = factor(genus, levels = genus_sizes$genus)) %>%
  dplyr::arrange(genus, virus_abbr) %>% dplyr::pull(name)

pa <- pa %>% dplyr::mutate(
  query         = factor(query, levels = isolate_order),
  ELMIdentifier = forcats::fct_rev(factor(ELMIdentifier)),
  genus         = factor(genus, levels = genus_sizes$genus))

ggplot(pa, aes(query, ELMIdentifier, fill = genus)) +
  geom_tile() +
  facet_grid(cols = vars(genus), scales = "free_x", space = "free_x") +
  scale_fill_viridis_d(option = "turbo", guide = "none") +
  labs(title = "ELM linear motifs across Potyviridae CI region fragments",
       x = "Isolate fragment (grouped by genus)", y = "ELM motif class") +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid = element_blank(), panel.spacing.x = unit(1.5, "pt"),
        strip.text.x = element_text(angle = 90, hjust = 0, size = 7),
        strip.background = element_rect(fill = "grey95", colour = NA))
```

## Genus-characteristic motifs

Motifs ranked by how much their prevalence varies across genera (restricted to
genera with ≥3 isolates). Because genus sizes are very unequal, each call
carries a **confidence** based on the size of the genus driving it, and a
**Wilson lower bound** on the prevalence — so a motif present in 3/3 isolates of
a small genus (point prevalence 1.0, lower bound ≈ 0.44) is not overstated
relative to one present in 166/166 *Potyvirus* isolates (lower bound ≈ 0.98).

```{r spec}
#| fig-width: 12
#| fig-height: 5.8
#| eval: !expr file.exists(file.path(here::here("data","elm_gget"),"elm_motif_specificity.tsv"))
spec <- readr::read_tsv(file.path(elm_dir, "elm_motif_specificity.tsv"),
                        show_col_types = FALSE)

MIN_N <- 3; CONF_LOW <- 5; TOP_M <- 30
big_genera <- genus_sizes %>% dplyr::filter(n_iso >= MIN_N) %>% dplyr::pull(genus)

top_motifs <- spec %>% dplyr::slice_max(spec_var, n = TOP_M) %>%
  dplyr::pull(ELMIdentifier)

prev_long <- spec %>%
  dplyr::filter(ELMIdentifier %in% top_motifs) %>%
  dplyr::select(ELMIdentifier, dplyr::all_of(as.character(big_genera))) %>%
  tidyr::pivot_longer(-ELMIdentifier, names_to = "genus", values_to = "prevalence")

lab_map <- genus_sizes %>% dplyr::filter(genus %in% big_genera) %>%
  dplyr::mutate(label = sprintf("%s (n=%d)%s", genus, n_iso,
                                ifelse(n_iso < CONF_LOW, " †", "")))
lm <- setNames(lab_map$label, lab_map$genus)

prev_long <- prev_long %>% dplyr::mutate(
  genus_lab = factor(lm[genus], levels = rev(lm[as.character(big_genera)])),
  ELMIdentifier = factor(ELMIdentifier,
    levels = spec$ELMIdentifier[spec$ELMIdentifier %in% top_motifs]))

lowconf <- spec$ELMIdentifier[spec$confidence == "low"]
prev_long <- prev_long %>%
  dplyr::mutate(lowconf = ELMIdentifier %in% lowconf & prevalence > 0.5)

ggplot(prev_long, aes(ELMIdentifier, genus_lab, fill = prevalence)) +
  geom_tile(colour = "grey92") +
  geom_tile(data = dplyr::filter(prev_long, lowconf),
            colour = "#2b8cbe", linewidth = 0.7, fill = NA) +
  scale_fill_viridis_c(option = "magma", direction = -1, limits = c(0, 1),
                       name = "Prevalence") +
  labs(title = "Genus-characteristic ELM motifs",
       caption = paste0("† genus with <", CONF_LOW,
                        " isolates (low-confidence); blue-ringed cells: ",
                        "characteristic call from a small genus."),
       x = "ELM motif class", y = "Genus") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        plot.caption = element_text(hjust = 0, size = 7, colour = "grey30"),
        panel.grid = element_blank())
```

The highest-confidence genus-characteristic calls (largest driving genus):

```{r spec-table}
#| eval: !expr file.exists(file.path(here::here("data","elm_gget"),"elm_motif_specificity.tsv"))
spec %>%
  dplyr::arrange(dplyr::desc(spec_var)) %>%
  dplyr::transmute(ELMIdentifier, top_genus, top_n,
                   top_prev = round(top_prev, 2),
                   wilson_lwr = round(top_prev_wilson_lwr, 2),
                   confidence, spec_var = round(spec_var, 3)) %>%
  dplyr::slice_head(n = 15) %>%
  knitr::kable(caption = paste("Top 15 motifs by cross-genus prevalence",
                               "variance, with small-sample confidence."))
```

## Notes

- The concatenated-submission workflow (`build_concat()` + junction-hit
  flagging) used for the original single ELM webserver query is retained in
  `potytools` but is no longer needed here, since `gget` lets us query each
  isolate directly.
- The gget ELM database footprint is ~3 MB (ELM-curated instances/classes);
  the larger DIAMOND binary lives inside the disposable pixi env, not in the
  project.
