---
title: "Potyvirus Protein Motifs and Interaction Networks"
subtitle: "Functional signature motifs across the polyprotein and the curated CI/HC-Pro/VPg interactome"
author: "chrissi m."
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    page-layout: full
execute:
  echo: true
  warning: false
  message: false
---

## Overview

The potyvirus genome is translated as a single polyprotein that is cleaved into
**~11 mature proteins**, each carrying short functional signature motifs whose
substitution changes transmissibility, silencing suppression, replication or
movement [@revers2015]. This chapter (i) catalogues those motifs with their
roles and primary references, (ii) locates the linear ones in the ICTV
Potyviridae alignment [@inouenagata2022], and (iii) draws the curated
protein–protein interaction network that centres on the cylindrical inclusion
(CI) helicase — the protein carrying the DNAzyme-like locus studied in the
earlier chapters.

```{r setup}
#| include: false
library(Biostrings)
library(tidyverse)
library(igraph)
library(visNetwork)
library(potytools)
```

## Signature motif catalogue

```{r motif-catalogue}
motifs <- tibble::tribble(
  ~protein,     ~motif,      ~pattern,       ~role,                                              ~ref,
  "P1",         "protease",  "—",            "Serine protease; host-range adaptation",           "revers2015",
  "HC-Pro",     "KITC",      "KITC",         "Aphid transmission (N-terminal)",                  "valli2018",
  "HC-Pro",     "FRNK",      "FR[NR]K",      "RNA-silencing suppression; symptom severity",      "valli2018",
  "P3N-PIPO",   "G2A6",      "GA{5,7}",      "Transcriptional slippage → PIPO; cell-to-cell movement", "chung2008",
  "CI",         "Walker A",  "G.{4}GK[ST]",  "ATPase P-loop of the RNA helicase",                "sorel2014",
  "CI",         "DECH box",  "DECH",         "Helicase motif II (Mg2+ coordination)",            "sorel2014",
  "6K2",        "TM anchor", "—",            "ER membrane anchor; replication vesicles",         "revers2015",
  "VPg",        "eIF4E-bind","—",            "Genome-linked protein; binds host eIF4E (recessive resistance)", "robaglia2006",
  "NIa-Pro",    "catalytic", "H.{31,35}D.{60,70}GXCG", "Main protease catalytic triad",          "revers2015",
  "NIb",        "GDD",       "GDD",          "RdRp catalytic motif (polymerase)",                "revers2015",
  "CP",         "DAG",       "DAG",          "Aphid transmission determinant (N-terminal)",      "atreya1991"
)

knitr::kable(motifs, caption = "Potyvirus mature-protein signature motifs, roles and primary references.")
```

The motifs are backed by: HC-Pro KITC/FRNK [@valli2018]; the P3N-PIPO
frameshift/movement locus [@chung2008]; the CI helicase Walker-A/DECH motifs and
its interactome [@sorel2014]; VPg–eIF4E and recessive resistance
[@robaglia2006]; the CP DAG aphid-transmission determinant [@atreya1991]; and the
overall polyprotein organisation [@revers2015].

## Locating linear motifs in the ICTV alignment

```{r motif-scan}
aln  <- readAAStringSet(here::here("data", "OPSR.Poty.Fig3.v16_align.txt"))
cons <- as.character(consensusString(aln, ambiguityMap = "X", threshold = 0.4))
cons_nogap <- gsub("-", "", cons)

# Linear motifs searchable directly on the consensus (regex)
scan_set <- dplyr::filter(motifs, pattern != "—")
hits <- scan_set %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    match_pos = list(as.integer(gregexpr(pattern, cons_nogap, perl = TRUE)[[1]])),
    n_hits    = sum(match_pos > 0),
    first_pos = ifelse(n_hits > 0, match_pos[1], NA_integer_)) %>%
  dplyr::ungroup() %>%
  dplyr::select(protein, motif, pattern, n_hits, first_pos)

knitr::kable(hits,
  caption = "Signature motifs located on the consensus of the ICTV Potyviridae alignment (gap-free coordinates).")
```

```{r motif-scan-plot}
#| fig-width: 9
#| fig-height: 3
found <- dplyr::filter(hits, n_hits > 0)
ggplot(found, aes(x = first_pos, y = reorder(paste(protein, motif), first_pos))) +
  geom_point(size = 3, colour = "#A23B72") +
  geom_segment(aes(xend = first_pos, yend = paste(protein, motif)), x = 0, colour = "grey80") +
  labs(title = "Approximate motif positions along the potyvirus polyprotein consensus",
       x = "Consensus position (aa, gap-free)", y = NULL) +
  cowplot::theme_cowplot()
```

## Protein–protein interaction network

A curated interaction network of the potyvirus proteins, centred on CI, drawn
from the CI interactome review [@sorel2014], the HC-Pro multitasking review
[@valli2018], the movement/PIPO literature [@chung2008] and VPg–host factor
interactions [@robaglia2006].

```{r ppi-network}
nodes <- tibble::tibble(
  id    = c("P1","HC-Pro","P3","P3N-PIPO","6K1","CI","6K2","VPg","NIa-Pro","NIb","CP","eIF4E"),
  group = c(rep("viral", 11), "host"))
nodes$label <- nodes$id
nodes$title <- ifelse(nodes$group == "host", "Host translation factor", "Viral protein")

edges <- tibble::tribble(
  ~from,       ~to,        ~process,          ~ref,
  "VPg",       "eIF4E",    "translation/resistance", "robaglia2006",
  "VPg",       "NIb",      "replication",     "revers2015",
  "NIa-Pro",   "NIb",      "replication",     "revers2015",
  "CI",        "HC-Pro",   "movement",        "sorel2014",
  "CI",        "CP",       "movement",        "sorel2014",
  "CI",        "P3N-PIPO", "movement",        "chung2008",
  "CI",        "VPg",      "replication",     "sorel2014",
  "HC-Pro",    "HC-Pro",   "self (dimer)",    "valli2018",
  "HC-Pro",    "CP",       "aphid transmission", "valli2018",
  "CP",        "CP",       "encapsidation",   "revers2015")

edges$color <- dplyr::recode(edges$process,
  "movement" = "#2E86AB", "replication" = "#D62828",
  "translation/resistance" = "#8E44AD", .default = "#7f8c8d")
edges$title <- paste0(edges$process, " [", edges$ref, "]")

visNetwork(nodes, edges, height = "600px", width = "100%") %>%
  visGroups(groupname = "viral", color = "#f5b041") %>%
  visGroups(groupname = "host",  color = "#82e0aa", shape = "diamond") %>%
  visEdges(smooth = FALSE) %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 42)
```

## CI as a hub

The CI helicase is the most connected viral node — bridging replication (with
VPg/NIb) and cell-to-cell movement (with CP, HC-Pro and P3N-PIPO) [@sorel2014].
It is also the protein carrying the conserved DNAzyme-like locus (Chapters on the
DNAzyme motif and HGT) and the ELM-rich CI region (Protein Motif Analysis),
making it the natural focal point where the sequence-level and network-level
observations in this book converge.
