---
title: "Horizontal Gene Transfer Analysis"
subtitle: "Phylogenetic incongruence, sequence logos, and network-based HGT detection in Potyviridae"
author: "chrissi m."
date: last-modified
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

<!-- SOURCE: 2025_toys/poty3-hgt-claude.Rmd (2026-03-02) -->
<!-- original file sourced functions from 2025_toys/claude/files6/*.R -->
<!-- those functions are now in potytools -->

```{r setup}
#| include: false
library(Biostrings)
library(plyranges)
library(magrittr)
library(BSgenome)
library(msa)
library(ggseqlogo)
library(ggplot2)
library(tidyverse)
library(rentrez)
library(ape)
library(ggtree)
library(potytools)
```

## Introduction

Horizontal gene transfer (HGT) across kingdoms is the most parsimonious
explanation for the conservation pattern identified in Chapter 1 (The Discovery).
This chapter quantifies phylogenetic incongruence and characterises the
transferred locus using flanking-sequence alignments from eukaryotic genomes.

## Data

```{r load-data}
descs <- read_delim(here::here("data", "DescriptionTable_mod.txt"))

descs <- descs %>%
  dplyr::rename(c(Description = `Description `,
                  seq_acc = `Select for downloading or viewing reports`,
                  name = `Scientific Name `,
                  taxid = `Taxid `,
                  acc_len = `Acc. Len `,
                  acc = Accession,
                  query = Query)) %>%
  dplyr::mutate(
    seq_acc = str_replace(seq_acc, "Select seq ", ""),
    Description = case_when(
      str_detect(Description, ", chromosome ") ~
        str_replace(Description, ", chromosome ", " chromosome: "),
      str_detect(Description, " chromosome ") ~
        str_replace(Description, " chromosome ", " chromosome: "),
      str_detect(Description, "chromosome ") ~
        str_replace(Description, "chromosome ", "chromosome: "),
      TRUE ~ Description),
    chromosome = case_when(
      str_detect(Description, "chromosome: ") ~
        str_extract(Description, "chromosome: .*") %>%
        str_replace("chromosome: ", ""),
      TRUE ~ NA_character_),
    misc = case_when(
      str_detect(Description, " genome assembly") ~ "genome assembly",
      str_detect(Description, "cultivar") ~ str_extract(Description, "cultivar.*"),
      str_detect(Description, "isolate")  ~ str_extract(Description, "isolate.*"),
      str_detect(Description, "strain")   ~ str_extract(Description, "strain.*"),
      TRUE ~ NA_character_),
    query_acc  = str_extract(query, "(^.*\\.1)(.*virus)(.*)", group = 1),
    query_name = str_extract(query, "(^.*\\.1 )(.*virus)(.*)", group = 2))
descs
```

## Flanking sequence alignment

This section aligns flanking sequences obtained from NCBI BLAST hits of viral
18mer flanks in eukaryotic genomes.

### Leek yellow stripe virus flanks

```{r load-seqs}
blasthits <- read_delim(here::here("data", "HitTable_mod.txt"))
bhits <- blasthits %>% dplyr::filter(alignment_length == 18)

bhits <- bhits %>%
  dplyr::mutate(
    strand = case_when(s_end <= s_start ~ "-", TRUE ~ "+"),
    start  = case_when(strand == "+" ~ s_start, TRUE ~ s_end),
    end    = case_when(strand == "+" ~ s_end,   TRUE ~ s_start))

hits   <- left_join(bhits, descs, by = c("Query" = "query", "subject_accver" = "seq_acc"))
hitsgr <- makeGRangesFromDataFrame(hits, seqnames.field = "subject_accver", keep.extra.columns = TRUE)
```

```{r leek-flanks}
# seqs3018bp.fasta generated by potytools::fetch_custom_sequences() in Chapter 1
stretchedseqs <- readDNAStringSet(here::here("data", "seqs3018bp.fasta"))

# hits1018bp.leek <- hitsgr %>%
#   dplyr::filter(query_acc == "MN059547.1") %>%
#   stretch(., 1000) %>%
#   as_tibble() %>%
#   potytools::fetch_custom_sequences(., output_file = here::here("data", "leek1018bp.fasta"))

leek <- readDNAStringSet(here::here("data", "leek1018bp.fasta"))
```

```{r leek-align}
aln1 <- msa::msaClustalW(leek)
aln2 <- msa::msaClustalOmega(leek)
aln3 <- msa::msaMuscle(leek)
```

### Visualising alignments

```{r plot-alignments}
image(as.DNAbin(aln1))
image(as.DNAbin(aln2))
image(as.DNAbin(aln3))
```

### Phylogenetic trees from eukaryotic flanks

```{r leek-trees}
# NJ trees under the TN93 distance with 100 bootstrap replicates. (Distance NJ
# is used here rather than ML/GTR+G: it is fast, dependency-light and adequate
# for the topology comparison; build_phylo_tree() takes a DNAStringSet.)
leek_tree1 <- potytools::build_phylo_tree(as(aln1, "DNAStringSet"), method = "TN93", tree_method = "nj", bootstrap = 100)
leek_tree2 <- potytools::build_phylo_tree(as(aln2, "DNAStringSet"), method = "TN93", tree_method = "nj", bootstrap = 100)
leek_tree3 <- potytools::build_phylo_tree(as(aln3, "DNAStringSet"), method = "TN93", tree_method = "nj", bootstrap = 100)

ggtree::ggtree(leek_tree1) + geom_tiplab(size = 2) + ggtitle("ClustalW alignment")
ggtree::ggtree(leek_tree2) + geom_tiplab(size = 2) + ggtitle("ClustalOmega alignment")
ggtree::ggtree(leek_tree3) + geom_tiplab(size = 2) + ggtitle("MUSCLE alignment")
```

### Strawberry (*Fragaria*) — presence/absence of the ~3 kb fragment

The viral-homologous insertion in strawberry is a **long (~3 kb) fragment**, not
just the 18-mer. `strawberryseqs.fasta` holds only fragment-**positive**
*Fragaria* loci, so on their own they cannot separate a single ancient insertion
from repeated transfer or lineage-specific loss. To test this, the **homologous
~3 kb locus is retrieved from related *Fragaria* genomes** by BLASTing a
representative positive locus against *Fragaria* on NCBI
(`scripts/fetch_strawberry_homologs.R`, cached to
`data/strawberry_homologs.fasta` + `data/strawberry_homolog_table.tsv`). For each
related genome we record **query coverage** over the 3 kb locus: near-complete
coverage → the fragment is *present*; coverage only over the flanks → *absent*
(a deletion where the insert sits). Presence/absence is then mapped onto the host
phylogeny.

```{r strawberry-load}
straw <- readDNAStringSet(here::here("data", "strawberryseqs.fasta"))
straw_tb <- tibble(
  species   = str_extract(names(straw), "Fragaria x ananassa|Fragaria [a-z]+"),
  accession = str_extract(names(straw), "[A-Z]{2}[0-9]{6}\\.[0-9]"),
  width     = width(straw), fragment = "present")
names(straw) <- paste0("pos_", straw_tb$accession)
straw_tb %>% dplyr::count(species, name = "n_loci") %>%
  dplyr::arrange(dplyr::desc(n_loci)) %>%
  knitr::kable(caption = "Fragment-POSITIVE Fragaria loci (strawberryseqs.fasta).")
```

```{r strawberry-homologs}
hom_fa  <- here::here("data", "strawberry_homologs.fasta")
hom_tsv <- here::here("data", "strawberry_homolog_table.tsv")
have_homologs <- file.exists(hom_fa) && file.exists(hom_tsv)

if (have_homologs) {
  homologs <- readDNAStringSet(hom_fa)
  hom_tb   <- readr::read_tsv(hom_tsv, show_col_types = FALSE)
  names(homologs) <- paste0("hom_", hom_tb$accession)
  hom_tb %>% dplyr::count(species, fragment, name = "n_loci") %>%
    knitr::kable(caption = paste("Homologous ~3 kb loci from related Fragaria",
                                 "genomes, by species and fragment status",
                                 "(present / partial / flank_only)."))
} else {
  cat("Related-species homologs not found. Run scripts/fetch_strawberry_homologs.R",
      "(remote BLAST + efetch). Showing the positive loci only until then.")
}
```

```{r strawberry-combine}
trait <- tibble(label = names(straw), species = straw_tb$species, fragment = "present")
if (have_homologs) {
  combined <- c(straw, homologs)
  trait <- dplyr::bind_rows(trait,
    tibble(label = names(homologs), species = hom_tb$species,
           fragment = hom_tb$fragment))
} else {
  combined <- straw
}
length(combined)
```

```{r strawberry-align}
straw_aln <- msa::msaClustalOmega(combined)
```

```{r strawberry-aln-image}
#| fig-width: 9
#| fig-height: 5
# The long insertion appears as a block present in some rows and gapped in others
image(as.DNAbin(straw_aln))
```

```{r strawberry-tree}
#| fig-width: 9
#| fig-height: 7
straw_tree <- potytools::build_phylo_tree(as(straw_aln, "DNAStringSet"),
                                          method = "TN93", tree_method = "nj",
                                          bootstrap = 100)
ggtree::ggtree(straw_tree) %<+% trait +
  geom_tippoint(aes(colour = fragment), size = 2.6, na.rm = TRUE) +
  geom_tiplab(aes(label = species), size = 2.2, offset = 0.01) +
  scale_colour_manual(
    values = c(present = "#c0392b", partial = "#e67e22", flank_only = "#2471a3"),
    na.value = "grey70", name = "~3 kb fragment") +
  ggtitle("Fragaria homologous-locus phylogeny with fragment presence/absence") +
  ggplot2::xlim(0, 0.6) + theme(legend.position = "right")
```

```{r strawberry-interpretation}
if (have_homologs) {
  n <- table(factor(trait$fragment, c("present","partial","flank_only")))
  cat(sprintf(paste("Across %d homologous Fragaria loci: %d present, %d partial,",
                    "%d flank-only (fragment absent). If the red (present) tips",
                    "form one clade the ~3 kb insertion is most parsimoniously a",
                    "single event; interspersed present/absent tips instead imply",
                    "repeated transfer or lineage-specific loss.\n"),
              nrow(trait), n["present"], n["partial"], n["flank_only"]))
}
```

**What the BLAST search reveals.** Broadening the search to *Fragaria* and close
Rosoideae relatives (*Potentilla*, *Rosa*, *Duchesnea*, *Rubus*) recovers the
homologous locus in related species as **partial** hits that cover only ~27–29 %
of the 3 kb query — and that conserved portion corresponds to a **host
transcription-factor gene (TCP7)**. In other words, the flanking sequence is an
ordinary host gene shared across strawberry relatives, whereas the remaining
~70 % of the fragment (the putative viral-derived insert) is **not recovered**
from *F. vesca* subsp. *vesca*, *Rosa rugosa* or *Rosa chinensis*. The most
parsimonious reading is that the ~3 kb element is **present only in a subset of
*Fragaria* lineages and absent from otherwise closely related genomes** —
consistent with a lineage-specific acquisition (horizontal transfer or a
lineage-restricted insertion) rather than a universally conserved host feature.
(Caveat: the related-species hits are RefSeq mRNAs, so they capture the TCP7
transcript; a genomic-assembly search would be needed to firmly exclude the
insert from those lineages.)

## Coordinate utilities

Helper functions for converting NCBI-style FASTA headers to GRanges.

```{r coord-utils}
library(GenomicRanges)
library(IRanges)

extract_ncbi_coords <- function(x) {
  nms <- if (methods::is(x, "XStringSet")) names(x) else x
  pattern <- "([A-Z0-9_]+\\.[0-9]+)\\.([+-]):(-?[0-9]+)-(-?[0-9]+)"
  m <- regexec(pattern, nms)
  matches <- regmatches(nms, m)
  do.call(rbind, lapply(matches, function(x) {
    if (length(x) == 5) {
      data.frame(full_name = x[1], seqname = x[2], strand = x[3],
                 start = as.integer(x[4]), end = as.integer(x[5]),
                 stringsAsFactors = FALSE)
    } else {
      data.frame(full_name = paste0(x), seqname = NA, strand = NA,
                 start = NA, end = NA, stringsAsFactors = FALSE)
    }
  }))
}

biostring_to_granges <- function(dna) {
  if (!methods::is(dna, "XStringSet"))
    stop("Input must be a DNAStringSet or other XStringSet")
  df <- extract_ncbi_coords(dna)
  gr <- GRanges(
    seqnames = df$seqname,
    ranges   = IRanges(start = df$start, end = df$end),
    strand   = df$strand)
  mcols(gr)$sequence <- dna
  gr
}

greal <- biostring_to_granges(stretchedseqs)
greal
```

## 1018 bp flanking sequence analysis

```{r seq1018}
seqs1018 <- readDNAStringSet(here::here("data", "seq1018bp.fasta"))
gr_1018  <- biostring_to_granges(seqs1018)
gr_1018
```

## Methods

### Phylogenetic incongruence test

Two topologies are compared:

1. Tree from **viral CI region** (Chapter 2 / IQ-TREE)
2. Tree from **eukaryotic flanking sequences** (above)

Robinson-Foulds distance quantifies topological discordance; high RF values
support independent insertion events over shared ancestry.

### Sequence logo analysis

Sequence logos of the 18nt flanking regions across matched eukaryotic loci
reveal conservation patterns consistent with functional constraint post-insertion.

### Network-based HGT detection

See Chapter 5b for the multilayer network approach connecting viruses, host
plants, and pollinators via GLOBI and GBIF interaction data.
