---
title: "Host-Virus Interaction Networks"
subtitle: "Species interaction networks from GBIF and GLOBI; potyvirus host range in ecological context"
author: "Christina Muedsam"
date: last-modified
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

<!-- SOURCE: 2025_toys/copilot_networks1.Rmd + ecoDBs.Rmd; BioNexus app v3.0    -->
<!-- Interactive companion app: shiny/bionexus/app.R (run with shiny::runApp())   -->

```{r setup}
#| include: false
library(igraph)
library(ggraph)
library(tidyverse)
library(visNetwork)
library(rotl)
```

## Introduction

To check whether interactions between species harbouring viral flanking sequences
could explain the distribution of homologous loci, I built multilayer networks
connecting:

1. **Species–host plant** ecological interactions (from EPPO / GLOBI)
2. **Shared-host** projections (species sharing a common host)
3. **Phylogenetic** relationships (Open Tree of Life)

See also the interactive [BioNexus Shiny app](../shiny/bionexus/) which queries
GBIF and GLOBI in real time.

## Part 1: Bipartite species–host network

```{r species-hosts}
species_hosts <- list(
  "Fallopia convolvulus" = character(0),
  "Luperina nickerlii"   = c("Elytrigia juncea", "Poaceae grasses"),
  "Luperina testacea"    = c("Poaceae grasses", "Dactylis glomerata"),
  "Trichiura crataegi"   = c("Prunus spinosa", "Crataegus", "Salix", "Carpinus"),
  "Sphenella marginata"  = c("Senecio/Jacobaea ragworts")
)

edges <- enframe(species_hosts, name = "species", value = "hosts") %>%
  unnest(hosts, keep_empty = TRUE)

host_nodes <- edges$hosts[!is.na(edges$hosts)]

bip_graph <- graph_from_data_frame(
  d = edges %>% filter(!is.na(hosts)),
  vertices = tibble(
    name = c(names(species_hosts), unique(host_nodes)),
    type = c(
      rep(FALSE, length(species_hosts)),
      rep(TRUE, length(unique(host_nodes)))
    )
  ),
  directed = FALSE
)

stopifnot(!any(is.na(V(bip_graph)$type)))
```

```{r plot-bipartite}
#| fig-cap: "Bipartite virus–host network: the two node classes (viral species, orange; host plants, blue) occupy the two rows, and an edge joins a virus to a plant it infects. Hubs are generalist viruses (or widely-infected hosts)."
ggraph(bip_graph, layout = "bipartite") +
  geom_edge_link() +
  geom_node_point(aes(color = type), size = 6) +
  geom_node_text(aes(label = name), repel = TRUE, size = 4) +
  scale_color_manual(values = c("FALSE" = "steelblue", "TRUE" = "darkorange"),
                     labels = c("species", "host plant")) +
  theme_void() +
  ggtitle("Species–Host Bipartite Network")
```

```{r species-projection}
#| fig-cap: "One-mode projection of the bipartite network onto the viral species: two viruses are linked if they share at least one host plant. Edges now represent ecological overlap (shared host range) rather than direct infection."
proj          <- bipartite_projection(bip_graph)
species_graph <- proj$proj1

ggraph(species_graph, layout = "fr") +
  geom_edge_link() +
  geom_node_point(color = "steelblue", size = 8) +
  geom_node_text(aes(label = name), repel = TRUE, size = 4) +
  theme_void() +
  ggtitle("Species-only Projected Network")
```

# Part 2: Multilayer network with phylogenetic and ecological edges

```{r multilayer-setup}
species <- names(species_hosts)

eco_edges <- enframe(species_hosts, name = "species", value = "hosts") %>%
  unnest(hosts, keep_empty = TRUE)
hosts <- unique(eco_edges$hosts[!is.na(eco_edges$hosts)])
```

```{r open-tree}
# Match species names to OTT IDs via the Open Tree of Life API (network).
# Wrapped so the chapter still renders offline / if the service is down —
# phylo_edges then falls back to empty and the multilayer network simply
# omits the phylogenetic layer.
phylo_edges <- tryCatch({
  taxa    <- rotl::tnrs_match_names(species)
  ott_ids <- taxa$ott_id[!is.na(taxa$ott_id)]
  tree    <- rotl::tol_induced_subtree(ott_ids = ott_ids)
  tibble(
    from = tree$tip.label[tree$edge[, 1]],
    to   = tree$tip.label[tree$edge[, 2]]
  ) %>%
    filter(!is.na(from), !is.na(to)) %>%
    mutate(type = "phylogenetic", weight = 1)
}, error = function(e) {
  message("Open Tree of Life query unavailable (", conditionMessage(e),
          "); omitting the phylogenetic layer.")
  tibble(from = character(), to = character(),
         type = character(), weight = numeric())
})
```

```{r shared-host-edges}
pair_grid <- expand.grid(sp1 = species, sp2 = species, stringsAsFactors = FALSE)

shared_host_edges <- pair_grid %>%
  filter(sp1 < sp2) %>%
  mutate(
    shared = map2_int(sp1, sp2, ~ length(intersect(
      species_hosts[[.x]], species_hosts[[.y]])))
  ) %>%
  filter(shared > 0) %>%
  transmute(from = sp1, to = sp2, type = "shared_host", weight = shared)
```

```{r combine-edges}
eco_edges2 <- eco_edges %>%
  filter(!is.na(hosts)) %>%
  transmute(from = species, to = hosts, type = "ecological", weight = 1)

all_edges <- bind_rows(eco_edges2, phylo_edges, shared_host_edges)
```

```{r node-traits}
traits <- tibble(
  name = c(species, hosts),
  trophic = case_when(
    name %in% species ~ "herbivore",
    name %in% hosts   ~ "plant",
    TRUE              ~ "unknown"),
  mobility = case_when(
    name %in% species ~ "flying",
    name %in% hosts   ~ "sessile",
    TRUE              ~ "unknown"),
  taxon = case_when(
    name %in% species ~ "Insecta",
    name %in% hosts   ~ "Plantae",
    TRUE              ~ "internal")
)
```

```{r visnetwork}
#| fig-cap: "Interactive multilayer virus–host–interaction network. Edge colour encodes the relationship type (orange/green/blue for the layers combined above); drag nodes to explore, hover for taxonomic details. Built from GBIF/GLOBI records plus Open Tree taxonomy."
nodes <- traits %>%
  transmute(id = name, label = name, group = trophic,
            mobility = mobility, taxon = taxon)

edges_vis <- all_edges %>%
  transmute(
    from, to,
    value = weight,
    title = paste(type, "weight =", weight),
    color = case_when(
      type == "ecological"   ~ "darkorange",
      type == "phylogenetic" ~ "forestgreen",
      type == "shared_host"  ~ "steelblue"))

visNetwork(nodes, edges_vis, height = "900px", width = "100%") %>%
  visEdges(smooth = FALSE) %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visPhysics(stabilization = TRUE) %>%
  visLayout(randomSeed = 123)
```

## BioNexus Shiny App

The BioNexus v3.0 Shiny application provides an interactive interface for
querying GBIF and GLOBI interaction data for arbitrary species lists and
constructing multilayer networks with multiplex participation coefficients.

The app lives at [`shiny/bionexus/app.R`](../shiny/bionexus/app.R). To launch:

```r
shiny::runApp(here::here("shiny", "bionexus"))
```

Key features:

- GBIF occurrence mapping for user-supplied species
- GLOBI biotic interaction retrieval
- Multilayer network construction (ecological + phylogenetic + shared-host layers)
- Multiplex participation coefficient for identifying bridge species
- Interactive visualisation via `visNetwork`

### A saved BioNexus network: *Zucchini yellow mosaic virus* host community

A live Shiny app cannot be embedded in a static document (an interactive
`shinylive`/WASM build is a future option). Instead, the network below is
**reconstructed from a saved BioNexus session** — the GLOBI-derived biotic
interaction network built around a *Zucchini yellow mosaic virus* (a cucurbit
potyvirus) host list. **Hover a node** for its taxonomy and a **PhyloPic
silhouette**, resolved to the most specific rank PhyloPic provides (species →
genus → family → order → class) so the outline matches the organism rather than
a generic class icon.

```{r bionexus-session}
#| eval: !expr file.exists(here::here("data","bionexus_zucchini95.rds"))
sess <- readRDS(here::here("data", "bionexus_zucchini95.rds"))

# base64-embed cached PhyloPic silhouettes (per taxon, most specific rank
# available) into hover tooltips — looked up by node name
phylodir <- here::here("data", "phylopic")
sil_b64 <- function(taxon) {
  if (is.na(taxon) || !nzchar(taxon)) return("")
  f <- file.path(phylodir, paste0(gsub("[^A-Za-z0-9]", "_", taxon), ".png"))
  if (!file.exists(f)) return("")
  sprintf('<img src="data:image/png;base64,%s" height="46" style="float:right;margin-left:8px;">',
          base64enc::base64encode(f))
}

# Keep ONLY directly recorded interactions as EDGES. The saved BioNexus session
# padded the network with heuristically inferred edges (source == "Heuristic",
# e.g. a beetle "pollinating" every plant in the community) and a dense
# statistical "cooccurrence" layer — neither represents an observed interaction,
# so both are dropped as edges. Every queried species is still shown as a NODE
# (including those with no recorded interaction) so the full community background
# is legible at a glance; isolated nodes simply have no GLOBI-recorded biotic link.
biotic <- sess$edges_df %>%
  dplyr::filter(layer != "cooccurrence", source == "GLOBI")

nodes <- sess$nodes_df %>%
  dplyr::mutate(
    label = name,
    group = functional_group,
    title = mapply(function(nm, k, cl, or, fa) paste0(
      "<div style='max-width:230px;font-size:12px'>", sil_b64(nm),
      "<b><i>", nm, "</i></b><br><span style='color:#666'>",
      k, " &middot; ", cl, "<br>", or, " &middot; ", fa, "</span></div>"),
      name, kingdom, class, order, family, SIMPLIFY = TRUE))

id_of <- stats::setNames(sess$nodes_df$id, sess$nodes_df$name)
edges <- biotic %>%
  dplyr::transmute(
    from = id_of[from_name], to = id_of[to_name],
    title = paste0(layer, " (", confidence, ")"),
    color = dplyr::recode(layer, pollination = "#e67e22", herbivory = "#27ae60",
                          parasitism = "#c0392b", predation = "#8e44ad",
                          .default = "#95a5a6")) %>%
  dplyr::filter(!is.na(from), !is.na(to))

visNetwork(nodes, edges, height = "620px", width = "100%") %>%
  visNodes(size = 12, font = list(size = 13),
           borderWidth = 1, color = list(border = "#ffffff")) %>%
  visEdges(smooth = FALSE, color = list(opacity = 0.45), width = 0.6) %>%
  # Live force-directed physics (spring/bounce) rather than a frozen igraph layout:
  # forceAtlas2 gives a lively but self-settling arrangement; drag a node to see it
  # spring back into place.
  visPhysics(solver = "forceAtlas2Based",
             forceAtlas2Based = list(gravitationalConstant = -60,
                                     centralGravity = 0.012,
                                     springLength = 90, springConstant = 0.08,
                                     damping = 0.4, avoidOverlap = 0.5),
             stabilization = list(enabled = TRUE, iterations = 200),
             maxVelocity = 30, minVelocity = 0.75) %>%
  visOptions(highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE),
             nodesIdSelection = TRUE, selectedBy = "group") %>%
  visLayout(randomSeed = 7) %>%
  visLegend(main = "Functional group", width = 0.18, position = "right")
```

```{r bionexus-caption}
#| echo: false
#| eval: !expr file.exists(here::here("data","bionexus_zucchini95.rds"))
lay <- sort(unique(biotic$layer))
cat(sprintf(paste0(
  "**Ecological interactions among taxa returned by a *Zucchini yellow mosaic ",
  "virus*-seeded GBIF + GLOBI query** (a saved BioNexus session). All %d queried ",
  "species are shown as nodes; only the %d **directly GLOBI-recorded** biotic ",
  "interactions (%s layers; edge colour = layer) are drawn as edges. Heuristically ",
  "inferred edges and the dense statistical co-occurrence layer that padded the ",
  "original session are **excluded** as they do not represent observed interactions, ",
  "so most species appear unconnected. That sparseness is itself informative: GLOBI ",
  "holds only a handful of directly recorded biotic links for this community, ",
  "consistent with the sporadic rather than continuous connectivity noted in ",
  "Chapter 1. Node colour = functional group; hover a node for its taxonomy and a ",
  "PhyloPic silhouette.\n"),
  nrow(nodes), nrow(edges), paste(lay, collapse = "/")))
```

Silhouettes are fetched once with the rOpenSci `rphylopic` package
(`scripts/fetch_phylopics.R`) and cached, so the chapter renders offline
[@gearty2023].

## Summary

The network analysis places the species harbouring viral-homologous loci within
a connected ecological network. Pollinators and their host plants — including
known potyvirus hosts — cluster together, consistent with the hypothesis that
viral flanking sequences were transferred horizontally via a shared biological
vector during co-infection events.
