---
title: "Host-Virus Interaction Networks"
subtitle: "Species interaction networks from GBIF and GLOBI; potyvirus host range in ecological context"
author: "chrissi m."
date: last-modified
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

<!-- SOURCE: 2025_toys/copilot_networks1.Rmd + ecoDBs.Rmd; BioNexus app v3.0    -->
<!-- Interactive companion app: shiny/bionexus/app.R (run with shiny::runApp())   -->

```{r setup}
#| include: false
library(igraph)
library(ggraph)
library(tidyverse)
library(visNetwork)
library(rotl)
```

## Introduction

To check whether interactions between species harbouring viral flanking sequences
could explain the distribution of homologous loci, I built multilayer networks
connecting:

1. **Species–host plant** ecological interactions (from EPPO / GLOBI)
2. **Shared-host** projections (species sharing a common host)
3. **Phylogenetic** relationships (Open Tree of Life)

See also the interactive [BioNexus Shiny app](../shiny/bionexus/) which queries
GBIF and GLOBI in real time.

## Part 1: Bipartite species–host network

```{r species-hosts}
species_hosts <- list(
  "Fallopia convolvulus" = character(0),
  "Luperina nickerlii"   = c("Elytrigia juncea", "Poaceae grasses"),
  "Luperina testacea"    = c("Poaceae grasses", "Dactylis glomerata"),
  "Trichiura crataegi"   = c("Prunus spinosa", "Crataegus", "Salix", "Carpinus"),
  "Sphenella marginata"  = c("Senecio/Jacobaea ragworts")
)

edges <- enframe(species_hosts, name = "species", value = "hosts") %>%
  unnest(hosts, keep_empty = TRUE)

host_nodes <- edges$hosts[!is.na(edges$hosts)]

bip_graph <- graph_from_data_frame(
  d = edges %>% filter(!is.na(hosts)),
  vertices = tibble(
    name = c(names(species_hosts), unique(host_nodes)),
    type = c(
      rep(FALSE, length(species_hosts)),
      rep(TRUE, length(unique(host_nodes)))
    )
  ),
  directed = FALSE
)

stopifnot(!any(is.na(V(bip_graph)$type)))
```

```{r plot-bipartite}
ggraph(bip_graph, layout = "bipartite") +
  geom_edge_link() +
  geom_node_point(aes(color = type), size = 6) +
  geom_node_text(aes(label = name), repel = TRUE, size = 4) +
  scale_color_manual(values = c("FALSE" = "steelblue", "TRUE" = "darkorange"),
                     labels = c("species", "host plant")) +
  theme_void() +
  ggtitle("Species–Host Bipartite Network")
```

```{r species-projection}
proj          <- bipartite_projection(bip_graph)
species_graph <- proj$proj1

ggraph(species_graph, layout = "fr") +
  geom_edge_link() +
  geom_node_point(color = "steelblue", size = 8) +
  geom_node_text(aes(label = name), repel = TRUE, size = 4) +
  theme_void() +
  ggtitle("Species-only Projected Network")
```

# Part 2: Multilayer network with phylogenetic and ecological edges

```{r multilayer-setup}
species <- names(species_hosts)

eco_edges <- enframe(species_hosts, name = "species", value = "hosts") %>%
  unnest(hosts, keep_empty = TRUE)
hosts <- unique(eco_edges$hosts[!is.na(eco_edges$hosts)])
```

```{r open-tree}
# Match species names to OTT IDs via the Open Tree of Life API (network).
# Wrapped so the chapter still renders offline / if the service is down —
# phylo_edges then falls back to empty and the multilayer network simply
# omits the phylogenetic layer.
phylo_edges <- tryCatch({
  taxa    <- rotl::tnrs_match_names(species)
  ott_ids <- taxa$ott_id[!is.na(taxa$ott_id)]
  tree    <- rotl::tol_induced_subtree(ott_ids = ott_ids)
  tibble(
    from = tree$tip.label[tree$edge[, 1]],
    to   = tree$tip.label[tree$edge[, 2]]
  ) %>%
    filter(!is.na(from), !is.na(to)) %>%
    mutate(type = "phylogenetic", weight = 1)
}, error = function(e) {
  message("Open Tree of Life query unavailable (", conditionMessage(e),
          "); omitting the phylogenetic layer.")
  tibble(from = character(), to = character(),
         type = character(), weight = numeric())
})
```

```{r shared-host-edges}
pair_grid <- expand.grid(sp1 = species, sp2 = species, stringsAsFactors = FALSE)

shared_host_edges <- pair_grid %>%
  filter(sp1 < sp2) %>%
  mutate(
    shared = map2_int(sp1, sp2, ~ length(intersect(
      species_hosts[[.x]], species_hosts[[.y]])))
  ) %>%
  filter(shared > 0) %>%
  transmute(from = sp1, to = sp2, type = "shared_host", weight = shared)
```

```{r combine-edges}
eco_edges2 <- eco_edges %>%
  filter(!is.na(hosts)) %>%
  transmute(from = species, to = hosts, type = "ecological", weight = 1)

all_edges <- bind_rows(eco_edges2, phylo_edges, shared_host_edges)
```

```{r node-traits}
traits <- tibble(
  name = c(species, hosts),
  trophic = case_when(
    name %in% species ~ "herbivore",
    name %in% hosts   ~ "plant",
    TRUE              ~ "unknown"),
  mobility = case_when(
    name %in% species ~ "flying",
    name %in% hosts   ~ "sessile",
    TRUE              ~ "unknown"),
  taxon = case_when(
    name %in% species ~ "Insecta",
    name %in% hosts   ~ "Plantae",
    TRUE              ~ "internal")
)
```

```{r visnetwork}
nodes <- traits %>%
  transmute(id = name, label = name, group = trophic,
            mobility = mobility, taxon = taxon)

edges_vis <- all_edges %>%
  transmute(
    from, to,
    value = weight,
    title = paste(type, "weight =", weight),
    color = case_when(
      type == "ecological"   ~ "darkorange",
      type == "phylogenetic" ~ "forestgreen",
      type == "shared_host"  ~ "steelblue"))

visNetwork(nodes, edges_vis, height = "900px", width = "100%") %>%
  visEdges(smooth = FALSE) %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visPhysics(stabilization = TRUE) %>%
  visLayout(randomSeed = 123)
```

## BioNexus Shiny App

The BioNexus v3.0 Shiny application provides an interactive interface for
querying GBIF and GLOBI interaction data for arbitrary species lists and
constructing multilayer networks with multiplex participation coefficients.

The app lives at [`shiny/bionexus/app.R`](../shiny/bionexus/app.R). To launch:

```r
shiny::runApp(here::here("shiny", "bionexus"))
```

Key features:

- GBIF occurrence mapping for user-supplied species
- GLOBI biotic interaction retrieval
- Multilayer network construction (ecological + phylogenetic + shared-host layers)
- Multiplex participation coefficient for identifying bridge species
- Interactive visualisation via `visNetwork`

### A saved BioNexus network: *Turnip mosaic virus* host community

A live Shiny app cannot be embedded in a static document (an interactive
`shinylive`/WASM build is a future option). Instead, the network below is
**reconstructed from a saved BioNexus session** — the GLOBI-derived biotic
interaction network for a *Turnip mosaic virus* host list. **Hover a node** for
its taxonomy and a **PhyloPic silhouette** of its class.

```{r bionexus-session}
#| eval: !expr file.exists(here::here("data","bionexus_turnip95.rds"))
#| fig-width: 10
sess <- readRDS(here::here("data", "bionexus_turnip95.rds"))

# base64-embed cached PhyloPic silhouettes (by class) into hover tooltips
phylodir <- here::here("data", "phylopic")
sil_b64 <- function(cls) {
  if (is.na(cls) || !nzchar(cls)) return("")
  f <- file.path(phylodir, paste0(gsub("[^A-Za-z0-9]", "_", cls), ".png"))
  if (!file.exists(f)) return("")
  sprintf('<img src="data:image/png;base64,%s" height="46" style="float:right;margin-left:8px;">',
          base64enc::base64encode(f))
}

nodes <- sess$nodes_df %>%
  dplyr::mutate(
    label = name,
    group = functional_group,
    title = mapply(function(nm, k, cl, or, fa) paste0(
      "<div style='max-width:230px;font-size:12px'>", sil_b64(cl),
      "<b><i>", nm, "</i></b><br><span style='color:#666'>",
      k, " &middot; ", cl, "<br>", or, " &middot; ", fa, "</span></div>"),
      name, kingdom, class, order, family, SIMPLIFY = TRUE))

id_of <- stats::setNames(sess$nodes_df$id, sess$nodes_df$name)
edges <- sess$edges_df %>%
  dplyr::transmute(
    from = id_of[from_name], to = id_of[to_name],
    title = paste0(layer, " (", confidence, ")"),
    color = dplyr::recode(layer, pollination = "#e67e22", herbivory = "#27ae60",
                          parasitism = "#c0392b", .default = "#95a5a6")) %>%
  dplyr::filter(!is.na(from), !is.na(to))

visNetwork(nodes, edges, height = "620px", width = "100%") %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE,
             selectedBy = "group") %>%
  visEdges(smooth = FALSE) %>%
  visLegend(main = "Functional group") %>%
  visPhysics(stabilization = TRUE) %>%
  visLayout(randomSeed = 7)
```

Silhouettes are fetched once with the rOpenSci `rphylopic` package
(`scripts/fetch_phylopics.R`) and cached, so the chapter renders offline
[@gearty2023].

## Summary

The network analysis places the species harbouring viral-homologous loci within
a connected ecological network. Pollinators and their host plants — including
known potyvirus hosts — cluster together, consistent with the hypothesis that
viral flanking sequences were transferred horizontally via a shared biological
vector during co-infection events.
