---
title: "Horizontal Gene Transfer Analysis"
subtitle: "Phylogenetic incongruence, sequence logos, and network-based HGT detection in Potyviridae"
author: "Christina Muedsam"
date: last-modified
execute:
  echo: true
  warning: false
  message: false
---

<!-- SOURCE: 2025_toys/poty3-hgt-claude.Rmd (2026-03-02) -->
<!-- original file sourced functions from 2025_toys/claude/files6/*.R -->
<!-- those functions are now in potytools -->

```{r setup}
#| include: false
library(Biostrings)
library(plyranges)
library(magrittr)
library(BSgenome)
library(msa)
library(ggseqlogo)
library(ggplot2)
library(tidyverse)
library(rentrez)
library(ape)
library(ggtree)
library(potytools)
theme_set(theme_manuscript())
```

## Introduction

Horizontal gene transfer (HGT) across kingdoms is the most parsimonious
explanation for the conservation pattern identified in Chapter 1 (The Discovery).
This chapter quantifies phylogenetic incongruence and characterises the
transferred locus using flanking-sequence alignments from eukaryotic genomes.

## Data

```{r load-data}
descs <- read_delim(here::here("data", "DescriptionTable_mod.txt"))

descs <- descs %>%
  dplyr::rename(c(Description = `Description `,
                  seq_acc = `Select for downloading or viewing reports`,
                  name = `Scientific Name `,
                  taxid = `Taxid `,
                  acc_len = `Acc. Len `,
                  acc = Accession,
                  query = Query)) %>%
  dplyr::mutate(
    seq_acc = str_replace(seq_acc, "Select seq ", ""),
    Description = case_when(
      str_detect(Description, ", chromosome ") ~
        str_replace(Description, ", chromosome ", " chromosome: "),
      str_detect(Description, " chromosome ") ~
        str_replace(Description, " chromosome ", " chromosome: "),
      str_detect(Description, "chromosome ") ~
        str_replace(Description, "chromosome ", "chromosome: "),
      TRUE ~ Description),
    chromosome = case_when(
      str_detect(Description, "chromosome: ") ~
        str_extract(Description, "chromosome: .*") %>%
        str_replace("chromosome: ", ""),
      TRUE ~ NA_character_),
    misc = case_when(
      str_detect(Description, " genome assembly") ~ "genome assembly",
      str_detect(Description, "cultivar") ~ str_extract(Description, "cultivar.*"),
      str_detect(Description, "isolate")  ~ str_extract(Description, "isolate.*"),
      str_detect(Description, "strain")   ~ str_extract(Description, "strain.*"),
      TRUE ~ NA_character_),
    query_acc  = str_extract(query, "(^.*\\.1)(.*virus)(.*)", group = 1),
    query_name = str_extract(query, "(^.*\\.1 )(.*virus)(.*)", group = 2))
descs
```

## Flanking sequence alignment

This section aligns flanking sequences obtained from NCBI BLAST hits of viral
18mer flanks in eukaryotic genomes.

### Leek yellow stripe virus flanks

```{r load-seqs}
blasthits <- read_delim(here::here("data", "HitTable_mod.txt"))
bhits <- blasthits %>% dplyr::filter(alignment_length == 18)

bhits <- bhits %>%
  dplyr::mutate(
    strand = case_when(s_end <= s_start ~ "-", TRUE ~ "+"),
    start  = case_when(strand == "+" ~ s_start, TRUE ~ s_end),
    end    = case_when(strand == "+" ~ s_end,   TRUE ~ s_start))

hits   <- left_join(bhits, descs, by = c("Query" = "query", "subject_accver" = "seq_acc"))
hitsgr <- makeGRangesFromDataFrame(hits, seqnames.field = "subject_accver", keep.extra.columns = TRUE)
```

```{r leek-flanks}
# seqs3018bp.fasta generated by potytools::fetch_custom_sequences() in Chapter 1
stretchedseqs <- readDNAStringSet(here::here("data", "seqs3018bp.fasta"))

# hits1018bp.leek <- hitsgr %>%
#   dplyr::filter(query_acc == "MN059547.1") %>%
#   stretch(., 1000) %>%
#   as_tibble() %>%
#   potytools::fetch_custom_sequences(., output_file = here::here("data", "leek1018bp.fasta"))

leek <- readDNAStringSet(here::here("data", "leek1018bp.fasta"))
```

```{r leek-align}
aln1 <- msa::msaClustalW(leek)
aln2 <- msa::msaClustalOmega(leek)
aln3 <- msa::msaMuscle(leek)
```

### Visualising alignments

```{r plot-alignments}
image(as.DNAbin(aln1))
image(as.DNAbin(aln2))
image(as.DNAbin(aln3))
```

### Phylogenetic trees from eukaryotic flanks

```{r leek-trees}
# NJ trees under the TN93 distance with 100 bootstrap replicates. (Distance NJ
# is used here rather than ML/GTR+G: it is fast, dependency-light and adequate
# for the topology comparison; build_phylo_tree() takes a DNAStringSet.)
leek_tree1 <- potytools::build_phylo_tree(as(aln1, "DNAStringSet"), method = "TN93", tree_method = "nj", bootstrap = 100)
leek_tree2 <- potytools::build_phylo_tree(as(aln2, "DNAStringSet"), method = "TN93", tree_method = "nj", bootstrap = 100)
leek_tree3 <- potytools::build_phylo_tree(as(aln3, "DNAStringSet"), method = "TN93", tree_method = "nj", bootstrap = 100)

tree_bg <- theme_manuscript_bg()
ggtree::ggtree(leek_tree1) + geom_tiplab(size = 2) + ggtitle("ClustalW alignment") + tree_bg
ggtree::ggtree(leek_tree2) + geom_tiplab(size = 2) + ggtitle("ClustalOmega alignment") + tree_bg
ggtree::ggtree(leek_tree3) + geom_tiplab(size = 2) + ggtitle("MUSCLE alignment") + tree_bg
```

### Strawberry (*Fragaria*) — presence/absence of the ~3 kb fragment

::: {.callout-note}
## Which virus, and is strawberry even a host?
The strawberry loci here were recovered by BLASTing the **DNAzyme-flanking
sequence of one specific isolate — NC_003377.1, *Maize dwarf mosaic virus*
(MDMV)** — against plant genomes. MDMV is a **maize/grass-infecting potyvirus**,
and **strawberry is *not* a documented host of MDMV**. So the *Fragaria* match is
a **cross-kingdom sequence homology to the viral flank**, *not* evidence that
MDMV infects strawberry. Read the section below as "a genomic fragment in
strawberry happens to match the MDMV DNAzyme-flank region", with the origin of
that match (ancient HGT vs coincidence vs a shared host gene) still open.
:::

The viral-homologous insertion in strawberry is a **long (~3 kb) fragment**, not
just the 18-mer. `strawberryseqs.fasta` holds only fragment-**positive**
*Fragaria* loci, so on their own they cannot separate a single ancient insertion
from repeated transfer or lineage-specific loss. To test this, the **homologous
~3 kb locus is retrieved from related *Fragaria* genomes** by BLASTing a
representative positive locus against *Fragaria* on NCBI
(`scripts/fetch_strawberry_homologs.R`, cached to
`data/strawberry_homologs.fasta` + `data/strawberry_homolog_table.tsv`). For each
related genome we record **query coverage** over the 3 kb locus: near-complete
coverage → the fragment is *present*; coverage only over the flanks → *absent*
(a deletion where the insert sits). Presence/absence is then mapped onto the host
phylogeny.

```{r strawberry-load}
straw <- readDNAStringSet(here::here("data", "strawberryseqs.fasta"))
straw_tb <- tibble(
  species   = str_extract(names(straw), "Fragaria x ananassa|Fragaria [a-z]+"),
  accession = str_extract(names(straw), "[A-Z]{2}[0-9]{6}\\.[0-9]"),
  width     = width(straw), fragment = "present")
names(straw) <- paste0("pos_", straw_tb$accession)
straw_tb %>% dplyr::count(species, name = "n_loci") %>%
  dplyr::arrange(dplyr::desc(n_loci)) %>%
  knitr::kable(caption = "Fragment-POSITIVE Fragaria loci (strawberryseqs.fasta).")
```

```{r strawberry-homologs}
hom_fa  <- here::here("data", "strawberry_homologs.fasta")
hom_tsv <- here::here("data", "strawberry_homolog_table.tsv")
have_homologs <- file.exists(hom_fa) && file.exists(hom_tsv)

if (have_homologs) {
  homologs <- readDNAStringSet(hom_fa)
  hom_tb   <- readr::read_tsv(hom_tsv, show_col_types = FALSE)
  names(homologs) <- paste0("hom_", hom_tb$accession)
  hom_tb %>% dplyr::count(species, fragment, name = "n_loci") %>%
    knitr::kable(caption = paste("Homologous ~3 kb loci from related Fragaria",
                                 "genomes, by species and fragment status",
                                 "(present / partial / flank_only)."))
} else {
  cat("Related-species homologs not found. Run scripts/fetch_strawberry_homologs.R",
      "(remote BLAST + efetch). Showing the positive loci only until then.")
}
```

```{r strawberry-combine}
trait <- tibble(label = names(straw), species = straw_tb$species, fragment = "present")
if (have_homologs) {
  combined <- c(straw, homologs)
  trait <- dplyr::bind_rows(trait,
    tibble(label = names(homologs), species = hom_tb$species,
           fragment = hom_tb$fragment))
} else {
  combined <- straw
}

# Exact-duplicate accessions (e.g. multiple GenBank deposits of the same
# Fragaria orientalis locus) collapse to a zero-branch-length split in the NJ
# tree below, rendering as an overlapping tip with no visible connecting
# branch. Drop them here, before alignment/tree-building, keeping one
# representative per unique sequence; trait/label bookkeeping above (and the
# fragment-positive loci count table) is untouched since it's about locus
# presence, not tree topology.
combined <- combined[!duplicated(as.character(combined))]
length(combined)
```

```{r strawberry-align}
straw_aln <- msa::msaClustalOmega(combined)
```

```{r strawberry-aln-image}
#| fig-width: 9
#| fig-height: 5
# The long insertion appears as a block present in some rows and gapped in others
image(as.DNAbin(straw_aln))
```

```{r strawberry-tree}
#| fig-width: 9
#| fig-height: 7
straw_tree <- potytools::build_phylo_tree(as(straw_aln, "DNAStringSet"),
                                          method = "TN93", tree_method = "nj",
                                          bootstrap = 100)
ggtree::ggtree(straw_tree) %<+% trait +
  geom_tippoint(aes(colour = fragment), size = 2.6, na.rm = TRUE) +
  geom_tiplab(aes(label = species), size = 2.2, offset = 0.01) +
  scale_colour_manual(
    values = c(present = "#c0392b", partial = "#e67e22", flank_only = "#2471a3"),
    na.value = "grey70", name = "~3 kb fragment") +
  ggtitle("Fragaria homologous-locus phylogeny with fragment presence/absence") +
  ggplot2::xlim(0, 0.6) +
  theme(legend.position = "right") +
  theme_manuscript_bg()
```

```{r strawberry-interpretation}
if (have_homologs) {
  n <- table(factor(trait$fragment, c("present","partial","flank_only")))
  cat(sprintf(paste("Across %d homologous Fragaria loci: %d present, %d partial,",
                    "%d flank-only (fragment absent). If the red (present) tips",
                    "form one clade the ~3 kb insertion is most parsimoniously a",
                    "single event; interspersed present/absent tips instead imply",
                    "repeated transfer or lineage-specific loss.\n"),
              nrow(trait), n["present"], n["partial"], n["flank_only"]))
}
```

**What the broadened BLAST reveals.** Searching the ~3 kb positive locus against
the whole **Rosaceae** (`scripts/strawberry_broad_blast.R`, cached to
`data/strawberry_broad_blast_hsps.tsv`; corroborated by a 250-HSP megablast of
82 *Fragaria*/Rosaceae subjects in
`data/strawberry_megablast_hittable.txt`) shows the homology is **confined to the
two ends of the query, with a central gap** — exactly the pattern expected if a
lineage-specific insert sits between two host-derived flanks:

- **5′ flank (~bp 1–670):** conserved genomic sequence, recovered from *Malus
  domestica* and *Fragaria* chromosomes (~82 % identity).
- **central ~1.4 kb (~bp 670–2130):** **no Rosaceae counterpart** — the candidate
  horizontally-acquired / lineage-specific segment.
- **3′ flank (~bp 2130–3018):** the host **TCP7 transcription-factor gene**,
  matching TCP7 (and its paralogues) across *Fragaria*, *Rosa*, *Prunus*,
  *Malus*, *Pyrus* and *Potentilla*.

The many partial (~28–35 % coverage) hits spread across numerous *Fragaria*
chromosomes are TCP-family paralogues matching only the 3′ TCP7 flank, not the
insert. The most parsimonious reading is therefore that the ~3 kb element is a
host locus (a genomic region fused to *TCP7*) carrying an insert **present only
in a subset of *Fragaria* lineages and absent from otherwise closely related
genomes** — consistent with a lineage-specific acquisition rather than a
universally conserved host feature.

### Localizing the insert within the 3 kb (sliding window)

Horizontal acquisition, if any, is expected to sit in a **sub-region** of the
3 kb, not throughout. Using the combined alignment, we scan along a
representative signature-positive locus and, in sliding windows, measure two
tracks: **conservation among the positive loci** and **identity to the TCP7
relatives**. The **host TCP7 region** should score high on *both*; a
**lineage-specific insert** should be conserved among positives but have *no*
counterpart in the relatives (the track drops to zero).

```{r strawberry-sliding}
#| fig-width: 10
#| fig-height: 4
mm      <- as.matrix(as(straw_aln, "DNAStringSet"))   # aligned char matrix
rn      <- rownames(mm)
pos_rows <- grep("^pos_", rn); hom_rows <- grep("^hom_", rn)
ref_row <- pos_rows[1]
refv    <- mm[ref_row, ]
ref_cols <- which(refv != "-")                        # coordinates along the ref locus

frac_match <- function(rows) {
  if (length(rows) == 0) return(rep(NA_real_, ncol(mm)))
  ref_mat <- matrix(refv, nrow = length(rows), ncol = length(refv), byrow = TRUE)
  colMeans(mm[rows, , drop = FALSE] == ref_mat)
}
wavg <- function(x, w = 90) as.numeric(stats::filter(x, rep(1 / w, w), sides = 2))

track <- tibble::tibble(
  pos = seq_along(ref_cols),
  `Conservation among positive loci`  = wavg(frac_match(setdiff(pos_rows, ref_row))[ref_cols]),
  `Identity to TCP7 relatives`        = wavg(frac_match(hom_rows)[ref_cols])) %>%
  tidyr::pivot_longer(-pos, names_to = "track", values_to = "value")

# shade where TCP7 homology is present (host-gene region)
host <- track %>% dplyr::filter(track == "Identity to TCP7 relatives", value > 0.3)
host_span <- if (nrow(host)) range(host$pos) else c(NA, NA)

ggplot(track, aes(pos, value, colour = track)) +
  { if (all(!is.na(host_span)))
      annotate("rect", xmin = host_span[1], xmax = host_span[2], ymin = 0, ymax = 1,
               alpha = 0.12, fill = "#2471a3") } +
  geom_line(linewidth = 0.8, na.rm = TRUE) +
  scale_colour_manual(values = c("Conservation among positive loci" = "#c0392b",
                                 "Identity to TCP7 relatives" = "#2471a3"), name = NULL) +
  labs(title = "Sliding-window localization of the strawberry insert",
       subtitle = "Blue shading = host TCP7 region (homology to relatives); the insert is conserved among positives but absent from relatives",
       x = "Position along representative positive locus (bp)", y = "Windowed identity") +
  ylim(0, 1) + theme_manuscript() + theme(legend.position = "top")
```

```{r strawberry-sliding-note}
if (all(!is.na(host_span))) {
  cat(sprintf(paste("The TCP7 host-gene homology maps to ~bp %d-%d of the locus;",
                    "the flanking sequence outside this window is conserved among",
                    "the signature-positive Fragaria loci yet has no counterpart in",
                    "the relatives — i.e. the candidate horizontally-acquired segment",
                    "is localized, not distributed across the whole 3 kb.\n"),
              host_span[1], host_span[2]))
}
```

#### BLAST hit distribution across the 3 kb locus

The sliding window above uses only the three RefSeq TCP7 mRNAs, which align to the
3′ flank alone. The genomic Rosaceae-wide BLAST gives the fuller picture: below,
each row is a hit subject and each segment is one HSP placed at its **query
coordinates**, coloured by subject type. The two homology blocks at the ends —
and the empty central band (yellow) — localise the insert directly, reproducing
the two-flank pattern seen in a full NCBI megablast (`data/…_megablast_hittable.txt`).

```{r strawberry-blast-distribution}
#| fig-width: 10
#| fig-height: 5.5
#| fig-cap: "Rosaceae-wide BLAST of the 3 kb positive locus. Each segment is one HSP at its query coordinates; rows are hit subjects. The 5′ block (~1–670 bp) is conserved genomic sequence (Malus/Fragaria chromosomes), the 3′ block (~2130–3018 bp) is the host TCP7 gene across Rosaceae, and the yellow central band (~670–2130 bp) has no Rosaceae hit — the candidate insert."
hsp <- readr::read_tsv(here::here("data", "strawberry_broad_blast_hsps.tsv"),
                       show_col_types = FALSE)
qlen <- 3018
hsp <- hsp %>% dplyr::mutate(
  kind = dplyr::case_when(
    grepl("TCP7", title) ~ "TCP7 gene (3′ flank)",
    grepl("chromosome|genome", title, ignore.case = TRUE) ~ "genomic (flank)",
    TRUE ~ "other"),
  # descriptive label: abbreviated organism + feature, accession kept in parens
  org = stringr::str_extract(sub("^PREDICTED: ", "", title), "^[A-Z][a-z]+ [a-z]+"),
  org = dplyr::coalesce(org, organism, "unknown"),
  org_abbr = sub("^([A-Z])[a-z]+ ", "\\1. ", org),
  feat = dplyr::case_when(
    grepl("pseudogene", title) ~ "TCP7 pseudogene",
    grepl("TCP7", title) ~ "TCP7 gene",
    grepl("chromosome", title, ignore.case = TRUE) ~
      sub("[Cc]hromosome:? *", "chr ", stringr::str_extract(title, "[Cc]hromosome:? *[0-9A-Za-z]+")),
    grepl("genome", title, ignore.case = TRUE) ~ "genome",
    TRUE ~ "locus"),
  label = sprintf("%s — %s  (%s)", org_abbr, feat, subject))
lev <- hsp %>% dplyr::group_by(label) %>%
  dplyr::summarise(m = min(qstart), .groups = "drop") %>%
  dplyr::arrange(m) %>% dplyr::pull(label)
hsp$label <- factor(hsp$label, levels = rev(lev))

ggplot(hsp) +
  annotate("rect", xmin = 670, xmax = 2130, ymin = -Inf, ymax = Inf,
           fill = "#f9e79f", alpha = 0.45) +
  geom_segment(aes(x = qstart, xend = qend, y = label, yend = label,
                   colour = kind, linewidth = pident)) +
  labs(subtitle = "yellow band = central insert (no Rosaceae hit)") +
  scale_linewidth(range = c(1, 3), guide = "none") +
  scale_colour_manual(values = c("TCP7 gene (3′ flank)" = "#2471a3",
                                 "genomic (flank)" = "#c0392b", "other" = "grey60"),
                      name = NULL) +
  scale_x_continuous(limits = c(0, qlen), expand = c(0.01, 0)) +
  labs(title = "Rosaceae BLAST of the 3 kb strawberry locus",
       x = "Position along the 3 kb query (bp)", y = NULL) +
  theme_manuscript() +
  theme(axis.text.y = element_text(size = 6), legend.position = "top")
```

```{r strawberry-blast-coverage}
#| fig-width: 10
#| fig-height: 2.6
#| fig-cap: "Per-position BLAST hit count across the 3 kb query. Two peaks (5′ genomic flank, 3′ TCP7 gene) bracket a central trough with zero hits — the localised candidate insert."
pos <- seq_len(qlen)
cnt <- colSums(outer(hsp$qstart, pos, "<=") & outer(hsp$qend, pos, ">="))
covdf <- data.frame(pos = pos, hits = cnt)

ggplot(covdf, aes(pos, hits)) +
  annotate("rect", xmin = 670, xmax = 2130, ymin = -Inf, ymax = Inf,
           fill = "#f9e79f", alpha = 0.45) +
  geom_area(fill = "#5d6d7e", alpha = 0.8) +
  labs(x = "Position along the 3 kb query (bp)", y = "BLAST hits") +
  theme_manuscript()
```

::: {.callout-tip collapse="true"}
## Proposed next steps (not yet run)
Working hypothesis: *if* there was any cross-kingdom transfer, it is most
parsimoniously a **single event propagated through strawberry breeding** (the
fragment appears in several *Fragaria* accessions). To test this properly:

1. **Delimit the insert precisely.** Pin down the exact boundaries of the central
   non-*TCP7* segment (the candidate insert), rather than the ~670–2130 bp
   estimate above.
2. **Re-BLAST the *whole* insert against viruses.** The initial search used only
   the 18-mer. If the fragment is genuinely MDMV-derived, the **full insert (not
   just the 18-mer) should also match the MDMV genome**; if it does not, a
   different (or no) viral origin is more likely.
3. **Fragment-internal phylogeny vs breeding history.** Build a tree from the
   large strawberry fragments alone and ask whether its topology tracks
   documented *Fragaria* breeding/domestication — a single-introduction-then-
   inheritance model predicts congruence.

These need live NCBI BLAST and curated breeding metadata, so they are recorded
here as next steps rather than executed.
:::

## Coordinate utilities

Helper functions for converting NCBI-style FASTA headers to GRanges.

```{r coord-utils}
library(GenomicRanges)
library(IRanges)

extract_ncbi_coords <- function(x) {
  nms <- if (methods::is(x, "XStringSet")) names(x) else x
  pattern <- "([A-Z0-9_]+\\.[0-9]+)\\.([+-]):(-?[0-9]+)-(-?[0-9]+)"
  m <- regexec(pattern, nms)
  matches <- regmatches(nms, m)
  do.call(rbind, lapply(matches, function(x) {
    if (length(x) == 5) {
      data.frame(full_name = x[1], seqname = x[2], strand = x[3],
                 start = as.integer(x[4]), end = as.integer(x[5]),
                 stringsAsFactors = FALSE)
    } else {
      data.frame(full_name = paste0(x), seqname = NA, strand = NA,
                 start = NA, end = NA, stringsAsFactors = FALSE)
    }
  }))
}

biostring_to_granges <- function(dna) {
  if (!methods::is(dna, "XStringSet"))
    stop("Input must be a DNAStringSet or other XStringSet")
  df <- extract_ncbi_coords(dna)
  gr <- GRanges(
    seqnames = df$seqname,
    ranges   = IRanges(start = df$start, end = df$end),
    strand   = df$strand)
  mcols(gr)$sequence <- dna
  gr
}

greal <- biostring_to_granges(stretchedseqs)
greal
```

## 1018 bp flanking sequence analysis

```{r seq1018}
seqs1018 <- readDNAStringSet(here::here("data", "seq1018bp.fasta"))
gr_1018  <- biostring_to_granges(seqs1018)
gr_1018
```

## Methods

### Phylogenetic incongruence test

Two topologies are compared:

1. Tree from **viral CI region** (Chapter 2 / IQ-TREE)
2. Tree from **eukaryotic flanking sequences** (above)

Robinson-Foulds distance quantifies topological discordance; high RF values
support independent insertion events over shared ancestry.

### Sequence logo analysis

Sequence logos of the 18nt flanking regions across matched eukaryotic loci
reveal conservation patterns consistent with functional constraint post-insertion.

### Network-based HGT detection

See Chapter 7 for the multilayer network approach connecting viruses, host
plants, and pollinators via GLOBI and GBIF interaction data.
