---
title: "Getting started with potytools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with potytools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

`potytools` provides a toolkit for analysing potyvirus genome sequences, with
emphasis on:

- GenBank file parsing (no Bioconductor/genbankr dependency)
- Codon usage bias (RSCU, ENC, CAI, GC content, wobble bias)
- Host classification (monocot vs. dicot)
- Motif-flanking region extraction and concatenation for short-linear-motif analysis
- NCBI sequence retrieval via `rentrez`

## Parsing GenBank files

```{r}
library(potytools)

# Parse all .gb files in a folder
records <- load_genbank_folder("path/to/genbank_files/")

# Get a flat data frame of all features
feat_df <- features_as_df(records)

# Extract all protein sequences
proteins <- get_all_proteins(records)
```

## Codon usage analysis

```{r}
library(Biostrings)

# Load CDS sequences
cds <- readDNAStringSet("ci_region_cds.fasta")

# RSCU across all sequences
rscu <- calculate_rscu(cds)

# Effective number of codons per sequence
enc_values <- sapply(as.character(cds), calculate_enc)

# GC content by codon position
gc <- calculate_gc_content(as.character(cds[[1]]))

# Compare motif region vs. full CDS
motifs <- readDNAStringSet("motif_18nt.fasta")
comparison <- compare_codon_usage(motifs, cds)
```

## Host classification

```{r}
seq_names <- names(cds)
host_table <- create_host_classification_table(seq_names)

# Assign host-specific codon tables (monocot or dicot)
host_codon_tables <- create_host_specific_codon_tables(host_table)

# Calculate CAI against each isolate's host
cai_df <- calculate_host_specific_cai(cds, host_codon_tables)
```

## Concatenating fragments for motif analysis

`build_concat()` joins named sequence fragments with a masking linker and
returns a position map, so downstream short-linear-motif hits can be traced
back to their source isolate. (The short-linear-motif search itself is run
externally with the `gget elm` Python tool; see the project book.)

```{r}
frags <- c(PVY = "MDFDYSKQ", TuMV = "LKPTGGVE", WMV = "AACDEFGA")
cc <- build_concat(frags)
cc$map
```

## Fetching custom sequences from NCBI

```{r}
coords <- data.frame(
  seqnames = c("MN908947", "NC_038311"),
  start    = c(1000L,      500L),
  end      = c(1200L,      700L),
  strand   = c("+",        "-")
)
fetch_custom_sequences(coords, output_file = "custom_seqs.fasta")
```
